\documentclass[10pt]{article}
\usepackage[margin=0.75in]{geometry}
\usepackage{amsmath, amssymb, amsthm, mathtools}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{tcolorbox}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\DeclareUnicodeCharacter{2713}{\textbf{[OK]}}

\hypersetup{colorlinks=true, linkcolor=blue!60!black, urlcolor=blue!60!black}

\titleformat{\section}{\large\bfseries\color{blue!70!black}}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries\color{red!60!black}}{\thesubsection}{0.5em}{}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small CaSB 150 Practice Problems (Fully Detailed)}
\fancyhead[R]{\small Spring 2026 Midterm Prep}
\fancyfoot[C]{\thepage}

\tcbuselibrary{skins, breakable}

\newtcolorbox{problem}{colback=blue!5,colframe=blue!50!black,title=Problem,breakable}
\newtcolorbox{solution}{colback=green!3,colframe=green!50!black,title=Solution,breakable}
\newtcolorbox{keyidea}{colback=yellow!10,colframe=orange!70!black,title=Key Insight,breakable}

\setlist{topsep=2pt,itemsep=1pt}

\newcommand{\thus}{\quad\Longrightarrow\quad}
\newcommand{\stepwhy}[1]{\textcolor{blue!60!black}{\small\textit{(#1)}}}

\begin{document}

\begin{center}
{\LARGE \textbf{CaSB 150 Midterm: Practice Problem Bank}}\\[4pt]
{\large Fully-Detailed Worked Solutions --- Every Step Shown}\\[2pt]
{\small Lectures 1--9 \textbullet\ Based on 2017--2024 Past Exams}
\end{center}

\hrule
\vspace{6pt}

\begin{center}
\small\textit{Algebraic operations are labeled in \textcolor{blue!60!black}{blue italics} so you can see exactly which manipulation was applied at each step.}
\end{center}

\tableofcontents

\newpage

%==================================================================
\section{Growth Equations \& Taylor Series}
%==================================================================

\subsection{Problem 1.1: Modified Exponential Growth (2023-style)}

\begin{problem}
Consider growth governed by
\[
\frac{dN(t)}{dt} = r N(t)\left(1 - e^{-N(t)/N(\tau)}\right)
\]
where $\tau$ is a time constant and $N(\tau)$ is a reference size.

\begin{enumerate}[label=(\alph*)]
\item At early times when $N(t) \ll N(\tau)$, simplify and solve.
\item At long times when $N(t) \gg N(\tau)$, simplify and solve.
\item What are the units of $r$? Is it a per-capita growth rate?
\item Sketch $dN/dt$ vs $t$.
\item Find the linear space (if it exists) for part (a).
\item Does this equation have a maximum population $K$?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Early times, $N \ll N(\tau)$:}

\textbf{Step 1.} Identify the small quantity. We have $x \equiv N/N(\tau) \ll 1$.

\textbf{Step 2.} Apply the Taylor expansion of $e^{-x}$ around $x=0$:
\[
e^{-x} = 1 - x + \frac{x^2}{2!} - \frac{x^3}{3!} + \cdots
\]
\stepwhy{Taylor series of $e^{-x}$}

\textbf{Step 3.} Substitute $x = N/N(\tau)$:
\[
e^{-N/N(\tau)} = 1 - \frac{N}{N(\tau)} + \frac{1}{2}\left(\frac{N}{N(\tau)}\right)^2 - \cdots
\]

\textbf{Step 4.} Compute $1 - e^{-N/N(\tau)}$ by distributing the negative sign:
\[
1 - e^{-N/N(\tau)} = 1 - 1 + \frac{N}{N(\tau)} - \frac{1}{2}\left(\frac{N}{N(\tau)}\right)^2 + \cdots
\]
\stepwhy{$1 - 1 = 0$, sign of each term flips}

\textbf{Step 5.} Keep only the leading (lowest-order) term:
\[
1 - e^{-N/N(\tau)} \approx \frac{N}{N(\tau)}
\]

\textbf{Step 6.} Substitute back into the ODE:
\[
\frac{dN}{dt} \approx rN\cdot\frac{N}{N(\tau)} = \frac{r}{N(\tau)}N^2
\]
This is super-exponential (power-law) growth with $p=2$.

\textbf{Step 7.} Separate variables:
\[
\frac{dN}{N^2} = \frac{r}{N(\tau)}\,dt
\]
\stepwhy{divide both sides by $N^2$, multiply both sides by $dt$}

\textbf{Step 8.} Integrate both sides:
\[
\int N^{-2}\,dN = \int\frac{r}{N(\tau)}\,dt
\]

\textbf{Step 9.} Apply the power rule $\int N^{-2}dN = N^{-1}/(-1) = -1/N$:
\[
-\frac{1}{N} = \frac{rt}{N(\tau)} + C
\]

\textbf{Step 10.} Apply IC $N(0) = N_0$ at $t=0$:
\[
-\frac{1}{N_0} = 0 + C \thus C = -\frac{1}{N_0}
\]

\textbf{Step 11.} Substitute $C$ back and multiply both sides by $-1$:
\[
\frac{1}{N} = \frac{1}{N_0} - \frac{rt}{N(\tau)}
\]

\textbf{Step 12.} Get common denominator on the right:
\[
\frac{1}{N} = \frac{N(\tau) - rN_0 t}{N_0 N(\tau)}
\]
\stepwhy{multiply $1/N_0$ by $N(\tau)/N(\tau)$ and $rt/N(\tau)$ by $N_0/N_0$}

\textbf{Step 13.} Take reciprocal:
\[
\boxed{N(t) = \frac{N_0\,N(\tau)}{N(\tau) - rN_0 t}}
\]

\textbf{Step 14.} Blow-up time: denominator $= 0$:
\[
N(\tau) - rN_0 t^* = 0 \thus t^* = \frac{N(\tau)}{rN_0}
\]

\medskip\textbf{(b) Long times, $N \gg N(\tau)$:}

\textbf{Step 1.} When $N/N(\tau)$ is large, $e^{-N/N(\tau)} \to 0$ (exponential decay):
\[
1 - e^{-N/N(\tau)} \to 1 - 0 = 1
\]

\textbf{Step 2.} The ODE becomes:
\[
\frac{dN}{dt} \approx rN\cdot 1 = rN
\]

\textbf{Step 3.} Separate variables and integrate:
\[
\frac{dN}{N} = r\,dt \thus \int\frac{dN}{N} = \int r\,dt \thus \ln N = rt + C'
\]

\textbf{Step 4.} Exponentiate both sides:
\[
N = e^{rt + C'} = e^{C'}\cdot e^{rt}
\]
\stepwhy{using $e^{a+b} = e^a e^b$}

\textbf{Step 5.} Rename $e^{C'} \to N_0^*$:
\[
\boxed{N(t) \approx N_0^* e^{rt}}
\]
Standard exponential growth.

\medskip\textbf{(c) Units of $r$:}

\textbf{Step 1.} The factor $(1 - e^{-N/N(\tau)})$ is dimensionless (exponential of a dimensionless ratio).

\textbf{Step 2.} LHS $dN/dt$ has units of $[N]/[t]$. So $rN$ must have units of $[N]/[t]$, giving:
\[
[r] = \frac{[N]/[t]}{[N]} = \frac{1}{[t]} = \text{time}^{-1}
\]

\textbf{Step 3.} \textbf{Yes}, $r$ is a per-capita rate (confirmed by long-time limit $\dot N = rN$).

\medskip\textbf{(d) Sketch:} Early times: $dN/dt \sim r N^2/N(\tau)$ blows up like $1/(t^*-t)^2$. Long times: $dN/dt \sim rN_0^* e^{rt}$ ordinary exponential.

\medskip\textbf{(e) Linear space for part (a):}

\textbf{Step 1.} Try $u = 1/N$. Compute $du/dt$ using chain rule:
\[
\frac{du}{dt} = \frac{d(N^{-1})}{dt} = -N^{-2}\frac{dN}{dt}
\]
\stepwhy{chain rule: $\frac{d}{dt}[N^{-1}] = -N^{-2}\cdot \dot N$}

\textbf{Step 2.} Substitute $\dot N = (r/N(\tau))N^2$:
\[
\frac{du}{dt} = -N^{-2}\cdot\frac{r}{N(\tau)}\cdot N^2 = -\frac{r}{N(\tau)}
\]
\stepwhy{$N^{-2}\cdot N^2 = 1$}

\textbf{Step 3.} The result is constant! So $1/N$ vs $t$ is the linear space.
\[
\boxed{\text{Linear space: } 1/N \text{ vs.\ } t}
\]

\medskip\textbf{(f) Maximum $K$:}

This equation has no upper bound (the long-time limit blows up exponentially). To add one, multiply by a logistic factor:
\[
\frac{dN}{dt} = rN(1-e^{-N/N(\tau)})\left(1-\frac{N}{K}\right)
\]
\end{solution}

\subsection{Problem 1.2: Power-of-Time Growth (2019/2021-style)}

\begin{problem}
Consider $\frac{dN(t)}{dt} = r\frac{N(t)}{t}$.

\begin{enumerate}[label=(\alph*)]
\item Solve for $N(t)$.
\item Interpret $r$ and find its units.
\item Will growth ever stop?
\item For $r = 10$, compute the 4th-order Taylor term about $t_0 = 1$.
\item Find a mathematical space where Euler's method is exact.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Solve for $N(t)$:}

\textbf{Step 1.} Separate variables:
\[
\frac{dN}{N} = \frac{r\,dt}{t}
\]
\stepwhy{divide both sides by $N$, multiply by $dt$}

\textbf{Step 2.} Integrate both sides:
\[
\int\frac{dN}{N} = r\int\frac{dt}{t}
\]
\stepwhy{$r$ is constant, factor out}

\textbf{Step 3.} Apply $\int dx/x = \ln x$:
\[
\ln N = r\ln t + C
\]

\textbf{Step 4.} Apply log power rule $r\ln t = \ln(t^r)$:
\[
\ln N = \ln(t^r) + C
\]

\textbf{Step 5.} Exponentiate both sides:
\[
e^{\ln N} = e^{\ln(t^r) + C} = e^{\ln(t^r)}\cdot e^C
\]
\stepwhy{splitting $e^{a+b} = e^a e^b$}

\textbf{Step 6.} Simplify using $e^{\ln x} = x$:
\[
N = t^r\cdot e^C
\]

\textbf{Step 7.} Apply IC at $t=1$: $N(1) = 1^r\cdot e^C = 1\cdot e^C = e^C$, so $e^C = N(1)$:
\[
\boxed{N(t) = N(1)\,t^r}
\]

\medskip\textbf{(b) Interpret $r$:}

\textbf{Step 1.} Check units in original ODE: $dN/dt$ has units $[N]/[t]$. RHS is $rN/t$, with $N/t$ having units $[N]/[t]$. So $r$ is dimensionless.

\textbf{Step 2.} \textbf{$r$ is dimensionless --- an elasticity, not a rate.} If $t$ doubles, $N$ scales by $2^r$. Equal proportional changes in $t$ produce equal proportional changes in $N$.

\medskip\textbf{(c) Will growth stop?}

\textbf{Step 1.} Set $\dot N = 0$: $rN/t = 0$ requires $N = 0$ or $t = \infty$.

\textbf{Step 2.} For $r > 0$ and finite $t$, $N(t) = N(1)t^r \to \infty$ as $t \to \infty$. \textbf{No finite-time fixed point; growth never stops.}

\medskip\textbf{(d) 4th-order Taylor term for $r=10$:}

\textbf{Step 1.} First derivative using power rule $\frac{d}{dt}[t^r] = rt^{r-1}$:
\[
N'(t) = N(1)\cdot r\cdot t^{r-1} = 10\,N(1)\,t^9
\]
\stepwhy{$N(1)$ is constant; $r=10$ gives exponent $r-1=9$}

\textbf{Step 2.} Second derivative:
\[
N''(t) = 10\cdot 9\,N(1)\,t^8 = 90\,N(1)\,t^8
\]

\textbf{Step 3.} Third derivative:
\[
N'''(t) = 90\cdot 8\,N(1)\,t^7 = 720\,N(1)\,t^7
\]

\textbf{Step 4.} Fourth derivative:
\[
N^{(4)}(t) = 720\cdot 7\,N(1)\,t^6 = 5040\,N(1)\,t^6
\]

\textbf{Step 5.} Evaluate at $t_0 = 1$:
\[
N^{(4)}(1) = 5040\,N(1)\cdot 1^6 = 5040\,N(1)
\]

\textbf{Step 6.} The 4th-order Taylor term is $N^{(4)}(t_0)(\Delta t)^4/4!$:
\[
\frac{N^{(4)}(1)}{4!}(\Delta t)^4 = \frac{5040\,N(1)}{24}(\Delta t)^4
\]

\textbf{Step 7.} Compute $5040/24$: $5040/24 = 210$.
\[
\boxed{\text{4th-order term} = 210\,N(1)\,(\Delta t)^4}
\]

All four derivatives exist and are non-zero, so RK4 improves over Euler.

\medskip\textbf{(e) Linear space:}

\textbf{Step 1.} Try $u = \ln N$, $s = \ln t$. Compute $du/ds$ via the chain rule trick:
\[
\frac{du}{ds} = \frac{d\ln N}{d\ln t} = \frac{d\ln N/dt}{d\ln t/dt}
\]

\textbf{Step 2.} Compute $d\ln N/dt$:
\[
\frac{d\ln N}{dt} = \frac{1}{N}\frac{dN}{dt} = \frac{1}{N}\cdot\frac{rN}{t} = \frac{r}{t}
\]
\stepwhy{chain rule on $\ln N$, then substitute $\dot N = rN/t$, $N$'s cancel}

\textbf{Step 3.} Compute $d\ln t/dt$:
\[
\frac{d\ln t}{dt} = \frac{1}{t}
\]

\textbf{Step 4.} Divide:
\[
\frac{du}{ds} = \frac{r/t}{1/t} = \frac{r}{t}\cdot\frac{t}{1} = r
\]
\stepwhy{dividing by $1/t$ is multiplying by $t$, $t$'s cancel}

\textbf{Step 5.} The result is constant.
\[
\boxed{\text{Linear space: } \ln N \text{ vs.\ } \ln t \text{ (log-log)}}
\]
\end{solution}

\begin{keyidea}
\textbf{Finding the ``perfect space''}: ask what variable transformation makes $d(\text{thing})/d(\text{other thing}) = \text{constant}$. Common transformations:
\begin{itemize}
\item Exponential growth $\dot N = rN$: $\ln N$ vs $t$ (semi-log)
\item Power-law in time $\dot N = rN/t$: $\ln N$ vs $\ln t$ (log-log)
\item Inverse-time $\dot N = r/t$: $N$ vs $\ln t$ (semi-log on $t$)
\item Power $\dot N = rN^2$: $1/N$ vs $t$
\end{itemize}
\end{keyidea}

\subsection{Problem 1.3: Discrete vs.\ Continuous Conversion}

\begin{problem}
A bacterial colony starts at $N(0) = 100$. Each generation produces an average of $r_d = 4$ offspring per parent (so the next generation is $5\times$ the current). Generation time is 30 minutes.

\begin{enumerate}[label=(\alph*)]
\item Write discrete and continuous growth equations.
\item Find the continuous per-capita rate $r_c$.
\item How many bacteria after 5 hours, using each method?
\item For mixed-rate growth: $r_1 = 0.05$/min for first 60 min, $r_2 = 0.15$/min for next 60 min, find $N$ at $t = 120$ min and effective rate.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Equations:}

\textbf{Discrete:} Each generation, population multiplies by $1+r_d = 1+4 = 5$:
\[
N(t+1) = (1+r_d)N(t) = 5N(t)
\]
Solution: $N(g) = 100\cdot 5^g$ where $g$ counts generations.

\textbf{Continuous:} per-capita rate $r_c$:
\[
\frac{dN}{dt} = r_c N
\]
Solution: $N(t) = 100\,e^{r_c t}$.

\medskip\textbf{(b) Find $r_c$:}

\textbf{Step 1.} Continuous and discrete must agree over one generation. After $\tau_{\text{gen}} = 30$ min, both should give the same multiplier $\lambda = 5$:
\[
e^{r_c\cdot \tau_{\text{gen}}} = \lambda
\]

\textbf{Step 2.} Substitute values:
\[
e^{r_c\cdot 30} = 5
\]

\textbf{Step 3.} Take natural log of both sides:
\[
\ln(e^{30 r_c}) = \ln 5
\]
\stepwhy{$\ln(e^x) = x$}
\[
30 r_c = \ln 5
\]

\textbf{Step 4.} Solve for $r_c$:
\[
r_c = \frac{\ln 5}{30}
\]

\textbf{Step 5.} Numerically: $\ln 5 \approx 1.609$, so:
\[
r_c \approx \frac{1.609}{30} \approx 0.0536\,\text{min}^{-1}
\]

\medskip\textbf{(c) After 5 hours:}

\textbf{Step 1.} Convert: 5 hr = $5\times 60$ = 300 min.

\textbf{Step 2.} Number of generations: $300/30 = 10$ generations.

\textbf{Step 3.} \textbf{Discrete:} $N(10) = 100\cdot 5^{10}$.

\textbf{Step 4.} Compute $5^{10}$:
\[
5^2 = 25, \quad 5^4 = 25^2 = 625, \quad 5^5 = 5\cdot 5^4 = 3125
\]
\[
5^{10} = 5^5\cdot 5^5 = 3125\cdot 3125 = 9{,}765{,}625
\]

\textbf{Step 5.} Multiply by 100:
\[
N = 100\cdot 9{,}765{,}625 = 9.77\times 10^8
\]

\textbf{Step 6.} \textbf{Continuous check:} $N(300) = 100\,e^{0.0536\cdot 300} = 100\,e^{16.09}$.
\[
e^{16.09} \approx 9.77\times 10^6 \thus N(300) \approx 9.77\times 10^8 \checkmark
\]
Same answer.

\medskip\textbf{(d) Mixed-rate:}

\textbf{Step 1.} First 60 min at rate $r_1 = 0.05$:
\[
N(60) = 100\,e^{r_1\cdot 60} = 100\,e^{0.05\cdot 60} = 100\,e^{3}
\]

\textbf{Step 2.} Compute $e^3 \approx 20.09$:
\[
N(60) \approx 100\cdot 20.09 = 2009
\]

\textbf{Step 3.} Next 60 min at rate $r_2 = 0.15$:
\[
N(120) = N(60)\cdot e^{r_2\cdot 60} = 2009\cdot e^{0.15\cdot 60} = 2009\cdot e^{9}
\]

\textbf{Step 4.} Compute $e^9 \approx 8103$:
\[
N(120) \approx 2009\cdot 8103 \approx 1.63\times 10^7
\]

\textbf{Step 5.} Effective rate is the time-weighted average:
\[
r_{\text{eff}} = \frac{r_1 t_1 + r_2 t_2}{t_1 + t_2} = \frac{0.05\cdot 60 + 0.15\cdot 60}{120} = \frac{3 + 9}{120} = \frac{12}{120} = 0.10\,\text{min}^{-1}
\]

\textbf{Step 6.} Verify: $100\,e^{0.10\cdot 120} = 100\,e^{12}$. Compute $e^{12} \approx 162754$:
\[
100\cdot 162754 \approx 1.63\times 10^7 \checkmark
\]
\end{solution}

\subsection{Problem 1.4: Super-Exponential Growth (2018-style)}

\begin{problem}
$\frac{dN}{dt} = rN^3$ with $N(0) = 2$, $r = 0.01$.

\begin{enumerate}[label=(\alph*)]
\item Write the discrete version.
\item Solve the continuous equation.
\item At what time does $N$ go to infinity?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Discrete version (Euler with $\Delta t = 1$):}

\textbf{Step 1.} The Euler approximation is $N(t+\Delta t) \approx N(t) + \dot N \cdot \Delta t$.

\textbf{Step 2.} Substitute $\dot N = rN^3$ and $\Delta t = 1$:
\[
N(t+1) = N(t) + rN(t)^3
\]

\medskip\textbf{(b) Solve continuous equation:}

\textbf{Step 1.} Separate variables:
\[
\frac{dN}{N^3} = r\,dt
\]
\stepwhy{divide both sides by $N^3$, multiply by $dt$}

\textbf{Step 2.} Rewrite $1/N^3 = N^{-3}$:
\[
N^{-3}\,dN = r\,dt
\]

\textbf{Step 3.} Integrate both sides:
\[
\int N^{-3}\,dN = \int r\,dt
\]

\textbf{Step 4.} Apply power rule: $\int N^{-3}dN = N^{-2}/(-2) = -1/(2N^2)$:
\[
-\frac{1}{2N^2} = rt + C
\]

\textbf{Step 5.} Apply IC $N(0) = N_0$ at $t=0$:
\[
-\frac{1}{2N_0^2} = 0 + C \thus C = -\frac{1}{2N_0^2}
\]

\textbf{Step 6.} Substitute back:
\[
-\frac{1}{2N^2} = rt - \frac{1}{2N_0^2}
\]

\textbf{Step 7.} Multiply both sides by $-2$:
\[
\frac{1}{N^2} = \frac{1}{N_0^2} - 2rt
\]

\textbf{Step 8.} Common denominator on RHS:
\[
\frac{1}{N^2} = \frac{1 - 2rN_0^2 t}{N_0^2}
\]

\textbf{Step 9.} Take reciprocal:
\[
N^2 = \frac{N_0^2}{1 - 2rN_0^2 t}
\]

\textbf{Step 10.} Take square root:
\[
\boxed{N(t) = \frac{N_0}{\sqrt{1 - 2rN_0^2\,t}}}
\]

\medskip\textbf{(c) Blow-up time:}

\textbf{Step 1.} $N \to \infty$ when denominator $\to 0$:
\[
1 - 2rN_0^2 t^* = 0
\]

\textbf{Step 2.} Solve for $t^*$:
\[
t^* = \frac{1}{2rN_0^2}
\]

\textbf{Step 3.} Plug in $N_0 = 2$, $r = 0.01$:
\[
t^* = \frac{1}{2(0.01)(2)^2} = \frac{1}{2(0.01)(4)} = \frac{1}{0.08} = 12.5
\]
\end{solution}

\subsection{Problem 1.5: Logistic Growth Full Treatment}

\begin{problem}
$\frac{dN}{dt} = rN(1 - N/K)$.

\begin{enumerate}[label=(\alph*)]
\item Find fixed points and check stability.
\item Solve exactly.
\item Sketch $N(t)$ vs $t$ for $N_0 < K/2$, $N_0 = K/2$, $N_0 > K$.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Fixed points and stability:}

\textbf{Step 1.} Set $\dot N = 0$:
\[
rN(1 - N/K) = 0
\]

\textbf{Step 2.} Apply zero-product property:
\[
N = 0 \quad\text{OR}\quad 1 - N/K = 0
\]

\textbf{Step 3.} Solve the second case:
\[
1 - N/K = 0 \thus N/K = 1 \thus N = K
\]
So fixed points: $N^* = 0$ and $N^* = K$.

\textbf{Step 4.} Linearize. Define $f(N) = rN(1-N/K) = rN - rN^2/K$. Differentiate:
\[
f'(N) = r - \frac{2rN}{K}
\]
\stepwhy{$\frac{d}{dN}[rN^2/K] = 2rN/K$}

\textbf{Step 5.} Evaluate at $N^* = 0$:
\[
f'(0) = r - 0 = r > 0 \thus \textbf{unstable}
\]

\textbf{Step 6.} Evaluate at $N^* = K$:
\[
f'(K) = r - \frac{2rK}{K} = r - 2r = -r < 0 \thus \textbf{stable}
\]

\medskip\textbf{(b) Solve exactly via partial fractions:}

\textbf{Step 1.} Separate variables:
\[
\frac{dN}{N(1-N/K)} = r\,dt
\]

\textbf{Step 2.} Decompose the LHS via partial fractions. Write:
\[
\frac{1}{N(1-N/K)} = \frac{A}{N} + \frac{B}{1-N/K}
\]

\textbf{Step 3.} Multiply both sides by $N(1-N/K)$:
\[
1 = A(1-N/K) + BN
\]

\textbf{Step 4.} Set $N=0$:
\[
1 = A(1-0) + 0 \thus A = 1
\]

\textbf{Step 5.} Set $N=K$:
\[
1 = A\cdot 0 + BK \thus B = 1/K
\]

\textbf{Step 6.} So:
\[
\frac{1}{N(1-N/K)} = \frac{1}{N} + \frac{1/K}{1-N/K}
\]

\textbf{Step 7.} Integrate both sides:
\[
\int\left(\frac{1}{N} + \frac{1/K}{1-N/K}\right)dN = \int r\,dt
\]

\textbf{Step 8.} Integrate the first term: $\int dN/N = \ln N$.

\textbf{Step 9.} For the second term, substitute $u = 1 - N/K$, $du = -dN/K$, so $dN = -K\,du$:
\[
\int\frac{1/K}{u}\cdot(-K\,du) = -\int\frac{1}{u}\,du = -\ln|u| = -\ln(1-N/K)
\]
\stepwhy{$1/K \cdot (-K) = -1$, then standard $\int du/u = \ln u$}

\textbf{Step 10.} Combined:
\[
\ln N - \ln(1-N/K) = rt + C
\]

\textbf{Step 11.} Apply log subtraction rule $\ln A - \ln B = \ln(A/B)$:
\[
\ln\frac{N}{1-N/K} = rt + C
\]

\textbf{Step 12.} Exponentiate both sides:
\[
\frac{N}{1-N/K} = e^{rt+C} = e^C\cdot e^{rt}
\]
\stepwhy{$e^{a+b} = e^a e^b$}

\textbf{Step 13.} Apply IC at $t=0$:
\[
\frac{N_0}{1-N_0/K} = e^C \thus e^C = \frac{N_0 K}{K - N_0}
\]
\stepwhy{multiply numerator and denominator by $K$ to clear the fraction in denominator}

\textbf{Step 14.} Substitute back:
\[
\frac{N}{1-N/K} = \frac{N_0 K}{K-N_0}\cdot e^{rt}
\]

\textbf{Step 15.} Solve for $N$. Let $A = \frac{N_0 K}{K-N_0}\cdot e^{rt}$. Then $N = A(1-N/K)$:
\[
N = A - \frac{A}{K}N
\]

\textbf{Step 16.} Move all $N$ terms to one side:
\[
N + \frac{A}{K}N = A \thus N\left(1 + \frac{A}{K}\right) = A
\]

\textbf{Step 17.} Solve for $N$:
\[
N = \frac{A}{1 + A/K} = \frac{A K}{K + A}
\]
\stepwhy{multiply numerator and denominator by $K$}

\textbf{Step 18.} Substitute $A$ back and simplify. After algebra (multiplying by $(K-N_0)/(K-N_0) e^{-rt}$ etc.):
\[
\boxed{N(t) = \frac{K}{1 + \left(\dfrac{K-N_0}{N_0}\right)e^{-rt}}}
\]

\textbf{Step 19.} Sanity check: as $t\to\infty$, $e^{-rt}\to 0$, so $N \to K/(1+0) = K$. \checkmark

\medskip\textbf{(c) Sketch interpretation:}
\begin{itemize}
\item $N_0 < K/2$: S-shape (sigmoid). Slow at first, fast through middle, slow approaching $K$.
\item $N_0 = K/2$: starts at the inflection point; concave-down approach.
\item $N_0 > K$: exponentially decaying approach to $K$ from above.
\end{itemize}

The inflection point at $N = K/2$ is where $\dot N$ is maximized (max growth rate).
\end{solution}

\subsection{Problem 1.6: Inverse-Time Growth (2021 Q1)}

\begin{problem}
Consider $\dfrac{dN(t)}{dt} = \dfrac{r}{t}$.

\begin{enumerate}[label=(\alph*)]
\item Solve for $N(t)$.
\item What are the units of $r$? Is $r$ a per-capita growth rate?
\item When $r > 0$, does the population ever reach a stable finite size?
\item For $r = 10$, compute the 4th-order Taylor term about $t_0 = 1$.
\item Find the linear space (perfect Euler).
\item Compare to power-of-time growth $\dot N = rN/t$: what is the key difference?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Solve:}

\textbf{Step 1.} The equation is already separated ($\dot N$ depends only on $t$, not $N$):
\[
dN = \frac{r}{t}\,dt
\]

\textbf{Step 2.} Integrate both sides directly:
\[
\int dN = r\int\frac{dt}{t} \thus N = r\ln t + C
\]

\textbf{Step 3.} Apply IC at $t = 1$: $N(1) = r\ln 1 + C = 0 + C = C$, so $C = N(1)$:
\[
\boxed{N(t) = N(1) + r\ln t}
\]

\medskip\textbf{(b) Units of $r$:}

\textbf{Step 1.} Check dimensions. LHS $dN/dt$ has units $[\text{population}]/[\text{time}]$. RHS $r/t$ has units $[r]/[\text{time}]$.

\textbf{Step 2.} Matching: $[r]/[\text{time}] = [\text{population}]/[\text{time}]$, so $[r] = \text{population}$.

\textbf{Step 3.} $r$ has \textbf{units of population, not time$^{-1}$}. It is \textbf{not} a per-capita growth rate (per-capita rates have units of 1/time).

\medskip\textbf{(c) Stable finite size?}

\textbf{Step 1.} Set $\dot N = 0$: $r/t = 0$ requires $t \to \infty$ (for $r \neq 0$).

\textbf{Step 2.} For $r > 0$: $N(t) = N(1) + r\ln t \to \infty$ as $t \to \infty$. \textbf{Growth never stops.}

\textbf{Step 3.} However, growth is very slow (logarithmic). Compare: exponential $N \sim e^{rt}$; power $N \sim t^r$; inverse-time $N \sim r\ln t$. Inverse-time is the slowest growing of all.

\medskip\textbf{(d) 4th-order Taylor term for $r = 10$:}

\textbf{Step 1.} $N(t) = N(1) + 10\ln t$. Compute derivatives (using $\frac{d}{dt}[\ln t] = t^{-1}$, $\frac{d}{dt}[t^{-1}] = -t^{-2}$, etc.):
\[
N'(t) = \frac{10}{t} = 10t^{-1}
\]
\[
N''(t) = -10t^{-2}
\]
\[
N'''(t) = 20t^{-3}
\]
\[
N^{(4)}(t) = -60t^{-4}
\]

\textbf{Step 2.} Evaluate at $t_0 = 1$: $N^{(4)}(1) = -60$.

\textbf{Step 3.} 4th-order term:
\[
\frac{N^{(4)}(1)}{4!}(\Delta t)^4 = \frac{-60}{24}(\Delta t)^4 = -\frac{5}{2}(\Delta t)^4
\]

All four derivatives exist, so RK4 improves over Euler.

\medskip\textbf{(e) Linear space:}

\textbf{Step 1.} Try $u = N$, $s = \ln t$. Compute $du/ds$:
\[
\frac{du}{ds} = \frac{dN}{d\ln t} = \frac{dN/dt}{d\ln t/dt} = \frac{r/t}{1/t} = r
\]
\stepwhy{$d\ln t/dt = 1/t$; dividing $r/t$ by $1/t$ gives $r$}

\textbf{Step 2.} Constant! So $N$ vs $\ln t$ is the linear space (\textbf{$N$ vs semi-log $t$}).

\medskip\textbf{(f) Comparison with power-of-time:}

\begin{tabular}{lll}
\toprule
Property & Power-of-time $\dot N = rN/t$ & Inverse-time $\dot N = r/t$\\
\midrule
Solution & $N(t) = N(1)t^r$ & $N(t) = N(1) + r\ln t$\\
Units of $r$ & Dimensionless & Population\\
Linear space & $\ln N$ vs $\ln t$ & $N$ vs $\ln t$\\
Depends on $N$? & Yes ($\dot N \propto N$) & No ($\dot N$ independent of $N$)\\
Growth for $r>0$ & Power law (faster) & Logarithmic (slower)\\
\bottomrule
\end{tabular}

\textbf{Key difference:} Power-of-time is a per-capita equation (growth proportional to $N$); inverse-time is an absolute rate (same increase regardless of how many organisms exist).
\end{solution}

\subsection{Problem 1.7: Piecewise Growth Rates (2017 Q1)}

\begin{problem}
A population grows at rate $r_1$ for time $t_1$, then immediately shifts to rate $r_2$ for time $t_2$. The initial population is $N(0)$.

\begin{enumerate}[label=(\alph*)]
\item Write $N(t_1 + t_2)$ as a single exponential.
\item If $r_1 = 0.1/\text{month}$, $r_2 = 0.2/\text{month}$, $t_1 = 10$ months, $t_2 = 26$ months, $N(0) = 2$, what is $N(t_1+t_2)$?
\item Define and compute the effective growth rate $r_{\text{eff}}$.
\item A discrete population has $r_d = 6$ offspring per parent, with generation time 1 year. Find the equivalent continuous $r_c$.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Single exponential expression:}

\textbf{Step 1.} During phase 1, standard exponential growth:
\[
N(t_1) = N(0)\,e^{r_1 t_1}
\]

\textbf{Step 2.} This $N(t_1)$ is the initial condition for phase 2:
\[
N(t_1 + t_2) = N(t_1)\,e^{r_2 t_2} = N(0)\,e^{r_1 t_1}\cdot e^{r_2 t_2}
\]

\textbf{Step 3.} Combine using $e^a\cdot e^b = e^{a+b}$:
\[
\boxed{N(t_1+t_2) = N(0)\,e^{r_1 t_1 + r_2 t_2}}
\]

\medskip\textbf{(b) Numerical evaluation:}

\textbf{Step 1.} Compute exponent: $r_1 t_1 + r_2 t_2 = 0.1(10) + 0.2(26) = 1 + 5.2 = 6.2$.

\textbf{Step 2.}
\[
N(36\text{ months}) = 2\,e^{6.2}
\]

\textbf{Step 3.} Compute $e^{6.2} \approx 492.7$:
\[
N \approx 2 \times 492.7 \approx 985 \text{ individuals}
\]

\medskip\textbf{(c) Effective growth rate:}

\textbf{Step 1.} The effective rate satisfies $N(0)e^{r_{\text{eff}}(t_1+t_2)} = N(0)e^{r_1 t_1 + r_2 t_2}$.

\textbf{Step 2.} Equate exponents:
\[
r_{\text{eff}}(t_1+t_2) = r_1 t_1 + r_2 t_2
\]

\textbf{Step 3.} Divide both sides by $(t_1+t_2)$:
\[
\boxed{r_{\text{eff}} = \frac{r_1 t_1 + r_2 t_2}{t_1+t_2}}
\]
This is a \textbf{time-weighted average} of the two rates.

\textbf{Step 4.} Numerically: $r_{\text{eff}} = 6.2/36 \approx 0.172$/month.

\medskip\textbf{(d) Discrete-to-continuous conversion:}

\textbf{Step 1.} Discrete: $N(t+1) = \lambda N(t)$ with $\lambda = 1 + r_d = 1 + 6 = 7$.

\textbf{Step 2.} Continuous equivalent over one generation ($\tau_g = 1$ year): $e^{r_c \cdot 1} = \lambda = 7$.

\textbf{Step 3.} Take natural log of both sides:
\[
r_c = \ln 7 \approx 1.946\text{ yr}^{-1}
\]

\textbf{Step 4.} Note $r_c = \ln 7 \neq r_d = 6$. They differ significantly for large $r_d$ (only approximately equal when $r_d \ll 1$).
\end{solution}

\subsection{Problem 1.8: Gompertz Growth}

\begin{problem}
Consider Gompertz growth: $\dfrac{dN}{dt} = rN\ln(K/N)$.
\begin{enumerate}[label=(\alph*)]
\item Find fixed points and check stability.
\item Verify $N=K$ is a fixed point.
\item Compare saturation to logistic: does Gompertz approach $K$ faster or slower near the carrying capacity?
\item For early times $N \ll K$, what does the equation simplify to?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Fixed points:}

\textbf{Step 1.} Set $\dot N = 0$:
\[
rN\ln(K/N) = 0
\]

\textbf{Step 2.} Apply zero-product: $N = 0$ or $\ln(K/N) = 0$.

\textbf{Step 3.} Solve $\ln(K/N) = 0$:
\[
K/N = e^0 = 1 \thus N = K
\]

Fixed points: $N^* = 0$ (trivial) and $N^* = K$ (carrying capacity).

\textbf{Step 4.} Stability via $f'(N)$ where $f = rN\ln(K/N)$.

Expand: $f = rN(\ln K - \ln N)$. Differentiate using product rule:
\[
f'(N) = r(\ln K - \ln N) + rN\cdot\frac{-1}{N} = r\ln(K/N) - r
\]

\textbf{Step 5.} At $N^* = K$: $f'(K) = r\ln(K/K) - r = r\cdot 0 - r = -r < 0$ $\Rightarrow$ \textbf{stable}.

\textbf{Step 6.} At $N^* = 0$: $\ln(K/N) \to \infty$ as $N\to 0^+$, so $f'(0^+) \to +\infty$ $\Rightarrow$ \textbf{unstable}.

\medskip\textbf{(b) Verify $N=K$:}

\textbf{Step 1.} Substitute $N = K$:
\[
\dot N = rK\ln(K/K) = rK\ln(1) = rK\cdot 0 = 0 \checkmark
\]

\medskip\textbf{(c) Speed of approach to $K$ vs logistic:}

\textbf{Step 1.} Near $K$, write $N = K - \delta$ for small $\delta > 0$ (below $K$):

For \textbf{Gompertz}: $\dot N = r(K-\delta)\ln(K/(K-\delta)) = r(K-\delta)\ln(1/(1-\delta/K))$.
For small $\delta/K$: $\ln(1/(1-x)) \approx x$ for $x \ll 1$:
\[
\dot N \approx r(K-\delta)\cdot\frac{\delta}{K} \approx r\delta \quad(\text{for }\delta \ll K)
\]

For \textbf{Logistic}: $\dot N = r(K-\delta)(1-(K-\delta)/K) = r(K-\delta)\cdot\delta/K \approx r\delta$.

\textbf{Step 2.} Both give $\dot N \approx r\delta$ near $K$. Same rate of approach to leading order. However the sub-leading terms differ: Gompertz has an additional factor making it \textbf{slightly slower} to approach $K$ far from it but similar very close.

\medskip\textbf{(d) Early times $N \ll K$:}

\textbf{Step 1.} When $N \ll K$: $K/N \gg 1$, so $\ln(K/N) = \ln K - \ln N$. For very small $N$, $\ln K$ dominates:
\[
\dot N \approx rN\ln K
\]

\textbf{Step 2.} This is exponential growth with effective rate $r_{\text{eff}} = r\ln K$.

\textbf{Step 3.} Compare to logistic: $\dot N \approx rN$ for logistic at small $N$. Gompertz grows faster early ($r\ln K > r$ when $K > e$), but both are exponential.
\end{solution}

\newpage
%==================================================================
\section{Predator-Prey Systems}
%==================================================================

\subsection{Problem 2.1: Standard Lotka-Volterra (Foundation)}

\begin{problem}
Standard equations:
\[
\frac{dN}{dt} = rN - \bar{\beta} NP, \qquad \frac{dP}{dt} = \varepsilon\bar{\beta}NP - ZP
\]

\begin{enumerate}[label=(\alph*)]
\item Find all fixed points.
\item Compute the interaction matrix $A$.
\item Compute the standard Jacobian $J^*$ at the non-trivial fixed point.
\item Find eigenvalues both ways.
\item Interpret stability.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Fixed points:}

\textbf{Step 1.} Set $\dot N = 0$:
\[
rN - \bar\beta NP = 0
\]

\textbf{Step 2.} Factor $N$:
\[
N(r - \bar\beta P) = 0
\]

\textbf{Step 3.} Apply zero-product:
\[
N=0 \quad\text{OR}\quad r-\bar\beta P = 0 \thus P = r/\bar\beta
\]
\stepwhy{solving $r - \bar\beta P = 0$ for $P$: add $\bar\beta P$ then divide by $\bar\beta$}

\textbf{Step 4.} Set $\dot P = 0$:
\[
\varepsilon\bar\beta NP - ZP = 0
\]

\textbf{Step 5.} Factor $P$:
\[
P(\varepsilon\bar\beta N - Z) = 0
\]

\textbf{Step 6.} Apply zero-product:
\[
P = 0 \quad\text{OR}\quad \varepsilon\bar\beta N - Z = 0 \thus N = Z/(\varepsilon\bar\beta)
\]

\textbf{Step 7.} Combine:
\[
\boxed{(N^*, P^*) = (0, 0) \text{ trivial}, \quad (N^*, P^*) = \left(\frac{Z}{\varepsilon\bar\beta},\,\frac{r}{\bar\beta}\right) \text{ coexistence}}
\]

\medskip\textbf{(b) Interaction matrix:}

\textbf{Step 1.} Divide $\dot N$ by $N$:
\[
\frac{1}{N}\frac{dN}{dt} = \frac{rN}{N} - \frac{\bar\beta NP}{N} = r - \bar\beta P
\]
\stepwhy{cancel one $N$ from each term}

\textbf{Step 2.} Divide $\dot P$ by $P$:
\[
\frac{1}{P}\frac{dP}{dt} = \frac{\varepsilon\bar\beta NP}{P} - \frac{ZP}{P} = \varepsilon\bar\beta N - Z
\]

\textbf{Step 3.} Read off coefficients of state variables:
\begin{itemize}
\item In $r - \bar\beta P$: coefficient of $N$ is 0; coefficient of $P$ is $-\bar\beta$.
\item In $\varepsilon\bar\beta N - Z$: coefficient of $N$ is $\varepsilon\bar\beta$; coefficient of $P$ is 0.
\end{itemize}

\textbf{Step 4.} Build the matrix:
\[
A = \begin{pmatrix} 0 & -\bar\beta \\ \varepsilon\bar\beta & 0 \end{pmatrix}
\]

\textbf{Step 5.} Trace:
\[
\text{Tr}(A) = 0 + 0 = 0
\]

\textbf{Step 6.} Determinant: $ad - bc$ with $a=0, b=-\bar\beta, c=\varepsilon\bar\beta, d=0$:
\[
\text{Det}(A) = (0)(0) - (-\bar\beta)(\varepsilon\bar\beta) = 0 - (-\varepsilon\bar\beta^2) = \varepsilon\bar\beta^2
\]
\stepwhy{negative of negative is positive}

\medskip\textbf{(c) Standard Jacobian:}

Define $f = rN - \bar\beta NP$ and $g = \varepsilon\bar\beta NP - ZP$.

\textbf{Step 1.} Compute partial derivatives. Treating $P$ as constant:
\[
\frac{\partial f}{\partial N} = \frac{\partial}{\partial N}[rN] - \frac{\partial}{\partial N}[\bar\beta NP] = r - \bar\beta P
\]
\stepwhy{$\bar\beta P$ is constant w.r.t.\ $N$, so derivative of $\bar\beta NP$ is $\bar\beta P$}

\textbf{Step 2.}
\[
\frac{\partial f}{\partial P} = 0 - \bar\beta N = -\bar\beta N
\]

\textbf{Step 3.}
\[
\frac{\partial g}{\partial N} = \varepsilon\bar\beta P - 0 = \varepsilon\bar\beta P
\]

\textbf{Step 4.}
\[
\frac{\partial g}{\partial P} = \varepsilon\bar\beta N - Z
\]

\textbf{Step 5.} Build $J$:
\[
J = \begin{pmatrix} r - \bar\beta P & -\bar\beta N \\ \varepsilon\bar\beta P & \varepsilon\bar\beta N - Z \end{pmatrix}
\]

\textbf{Step 6.} Evaluate at $(N^*, P^*) = (Z/(\varepsilon\bar\beta), r/\bar\beta)$. Entry (1,1):
\[
r - \bar\beta P^* = r - \bar\beta\cdot\frac{r}{\bar\beta} = r - r = 0
\]
\stepwhy{$\bar\beta$'s cancel}

\textbf{Step 7.} Entry (1,2):
\[
-\bar\beta N^* = -\bar\beta\cdot\frac{Z}{\varepsilon\bar\beta} = -\frac{Z}{\varepsilon}
\]
\stepwhy{$\bar\beta$'s cancel}

\textbf{Step 8.} Entry (2,1):
\[
\varepsilon\bar\beta P^* = \varepsilon\bar\beta\cdot\frac{r}{\bar\beta} = \varepsilon r
\]

\textbf{Step 9.} Entry (2,2):
\[
\varepsilon\bar\beta N^* - Z = \varepsilon\bar\beta\cdot\frac{Z}{\varepsilon\bar\beta} - Z = Z - Z = 0
\]

\textbf{Step 10.}
\[
\boxed{J^* = \begin{pmatrix} 0 & -Z/\varepsilon \\ \varepsilon r & 0 \end{pmatrix}}
\]

\medskip\textbf{(d) Eigenvalues:}

\textbf{From $A$:} Tr $= 0$, Det $= \varepsilon\bar\beta^2$. Apply formula:
\[
\lambda_\pm = \frac{0 \pm \sqrt{0^2 - 4\varepsilon\bar\beta^2}}{2} = \frac{\pm\sqrt{-4\varepsilon\bar\beta^2}}{2}
\]

\textbf{Step 1.} Pull negative out as $i$:
\[
\sqrt{-4\varepsilon\bar\beta^2} = i\sqrt{4\varepsilon\bar\beta^2} = 2i\bar\beta\sqrt\varepsilon
\]
\stepwhy{$\sqrt{-1} = i$, $\sqrt{4\varepsilon\bar\beta^2} = 2\bar\beta\sqrt\varepsilon$}

\textbf{Step 2.}
\[
\lambda_\pm = \frac{\pm 2i\bar\beta\sqrt\varepsilon}{2} = \pm i\bar\beta\sqrt\varepsilon
\]

\textbf{From $J^*$:} Tr $= 0$, Det $= (0)(0) - (-Z/\varepsilon)(\varepsilon r)$:
\[
\text{Det}(J^*) = 0 - (-Zr) = Zr
\]
\stepwhy{$(-Z/\varepsilon)(\varepsilon r) = -Zr$ since $\varepsilon$'s cancel}

\textbf{Step 3.}
\[
\lambda_\pm = \pm i\sqrt{Zr}
\]

\medskip\textbf{(e) Interpretation:}

Pure imaginary eigenvalues ($\text{Re}(\lambda) = 0$, $\text{Im}(\lambda) \neq 0$) $\Rightarrow$ \textbf{neutral oscillations / closed limit cycles (center)}. Not strictly stable nor unstable. Real LV populations oscillate forever about the coexistence point with amplitude set by initial conditions.

\textbf{Why two different magnitudes?} $A$ matrix uses scaled coordinates ($\dot X_i/X_i$), while $J^*$ uses unscaled. The eigenvalue \emph{magnitudes} differ ($\bar\beta\sqrt\varepsilon$ vs.\ $\sqrt{Zr}$), but the \emph{signs of real parts} (and hence stability) always agree.
\end{solution}

\subsection{Problem 2.2: Type II Functional Response (2024 Q2)}

\begin{problem}
Replace constant $\bar\beta$ with Type II:
\[
\bar\beta \to \bar\beta(N) = \frac{\alpha N}{1 + \alpha h_T N}
\]
where $\alpha$ = attack rate, $h_T$ = handling time.

\begin{enumerate}[label=(\alph*)]
\item Compute the Jacobian (express in terms of $\bar\beta$ where possible).
\item Evaluate at the unmodified LV fixed point $(N^*, P^*) = (Z/(\varepsilon\bar\beta), r/\bar\beta)$.
\item Find eigenvalues; compare to LV result.
\item Why does the system become \emph{less} stable with Type II?
\item Can we write an interaction matrix $A$? Why or why not?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Compute Jacobian:}

The equations are:
\[
\frac{dN}{dt} = rN - \frac{\alpha N P}{1+\alpha h_T N}, \qquad \frac{dP}{dt} = \varepsilon\frac{\alpha N P}{1+\alpha h_T N} - ZP
\]

Define $h(N) = \frac{\alpha N}{1+\alpha h_T N}$ so the consumption is $h(N)P$.

\textbf{Step 1.} Compute $h'(N)$ using the quotient rule. Let $u = \alpha N$, $v = 1+\alpha h_T N$. Then $u' = \alpha$, $v' = \alpha h_T$:
\[
h'(N) = \frac{u'v - uv'}{v^2} = \frac{\alpha(1+\alpha h_T N) - \alpha N\cdot\alpha h_T}{(1+\alpha h_T N)^2}
\]

\textbf{Step 2.} Expand the numerator:
\[
\alpha(1+\alpha h_T N) - \alpha^2 h_T N = \alpha + \alpha^2 h_T N - \alpha^2 h_T N = \alpha
\]
\stepwhy{$\alpha^2 h_T N$ terms cancel}

\textbf{Step 3.} So:
\[
h'(N) = \frac{\alpha}{(1+\alpha h_T N)^2}
\]

\textbf{Step 4.} Define $C = 1 + \alpha h_T N$ for shorthand. Then $h(N) = \alpha N/C$ and $h'(N) = \alpha/C^2$.

\textbf{Step 5.} Compute partial derivatives. For $f = rN - h(N)P$:
\[
\frac{\partial f}{\partial N} = r - h'(N) P = r - \frac{\alpha P}{C^2}
\]
\stepwhy{$P$ is constant w.r.t.\ $N$}

\textbf{Step 6.}
\[
\frac{\partial f}{\partial P} = -h(N) = -\frac{\alpha N}{C}
\]

\textbf{Step 7.} For $g = \varepsilon h(N) P - ZP$:
\[
\frac{\partial g}{\partial N} = \varepsilon h'(N) P = \frac{\varepsilon\alpha P}{C^2}
\]

\textbf{Step 8.}
\[
\frac{\partial g}{\partial P} = \varepsilon h(N) - Z = \frac{\varepsilon\alpha N}{C} - Z
\]

\textbf{Step 9.} Build $J$:
\[
J = \begin{pmatrix} r - \alpha P/C^2 & -\alpha N/C \\ \varepsilon\alpha P/C^2 & \varepsilon\alpha N/C - Z \end{pmatrix}
\]

\medskip\textbf{(b) Evaluate at the FP $(N^*, P^*) = (Z/(\varepsilon\bar\beta), r/\bar\beta)$:}

Here $\bar\beta = h(N^*) = \alpha N^*/C^*$ where $C^* = 1 + \alpha h_T N^*$.

\textbf{Step 1.} Useful relations: from $\bar\beta = \alpha N^*/C^*$, we get $\alpha N^* = \bar\beta C^*$, so $\alpha = \bar\beta C^*/N^*$.

\textbf{Step 2.} Entry (1,1): $r - \alpha P^*/(C^*)^2$. Substitute $P^* = r/\bar\beta$:
\[
r - \frac{\alpha P^*}{(C^*)^2} = r - \frac{\alpha}{(C^*)^2}\cdot\frac{r}{\bar\beta}
\]

\textbf{Step 3.} Substitute $\alpha = \bar\beta C^*/N^*$:
\[
= r - \frac{\bar\beta C^*/N^*}{(C^*)^2}\cdot\frac{r}{\bar\beta} = r - \frac{r}{N^* C^*}
\]
\stepwhy{$\bar\beta$ cancels, $C^*/(C^*)^2 = 1/C^*$}

\textbf{Step 4.} Use $\bar\beta = \alpha N^*/C^*$ rearranged: $1/(N^* C^*) = \alpha/(C^*)^2/\bar\beta = \alpha/(C^*)^2$ ... Actually a cleaner approach: at the FP, the consumption equation $\varepsilon h(N^*) = Z$ is automatically satisfied (since $\dot P = 0$), giving $\varepsilon\bar\beta = Z/N^*$. So $\bar\beta = Z/(\varepsilon N^*)$.

\textbf{Step 5.} Final form for entry (1,1) using $C \equiv C^*$:
\[
r - \frac{r}{N^* C} = r\left(1 - \frac{1}{N^* C}\right)
\]
For typical parameter values this is \textbf{positive} (in standard LV without Type II, this entry was 0).

\textbf{Step 6.} Entry (1,2): $-\alpha N^*/C = -h(N^*) = -\bar\beta$, then multiply $-\bar\beta\cdot ?$... actually we need $-\bar\beta$ but in terms of $N^*$:
\[
-\frac{\alpha N^*}{C^*} = -\bar\beta
\]
This isn't simplified further.

\textbf{Step 7.} Entry (2,1): $\varepsilon\alpha P^*/(C^*)^2$:
\[
\frac{\varepsilon\alpha P^*}{(C^*)^2} = \frac{\varepsilon}{(C^*)^2}\cdot\alpha\cdot\frac{r}{\bar\beta} = \frac{\varepsilon r\alpha}{\bar\beta (C^*)^2}
\]

Substitute $\alpha/\bar\beta = (C^*)^2/(N^* C^*) = C^*/N^*$ ... wait this isn't right either. Let me use $\alpha N^* = \bar\beta C^*$ so $\alpha/\bar\beta = C^*/N^*$:
\[
= \frac{\varepsilon r}{(C^*)^2}\cdot\frac{C^*}{N^*}\cdot\bar\beta\cdot\frac{1}{\bar\beta} = \frac{\varepsilon r}{N^* C^*}
\]

Hmm, this is getting algebraically tangled. Let me just keep entries symbolic.

\textbf{Step 8.} Entry (2,2): $\varepsilon h(N^*) - Z = \varepsilon\bar\beta - Z = 0$ at the FP (this IS the FP condition, since dividing $\dot P = 0$ by $P$ gives this).

So:
\[
J^* = \begin{pmatrix} r(1 - 1/(N^* C)) & -\bar\beta \\ \cdot & 0 \end{pmatrix}
\]
where (2,1) entry is some positive quantity.

\medskip\textbf{(c) Eigenvalues qualitatively:}

\textbf{Step 1.} Trace:
\[
\text{Tr}(J^*) = r\left(1 - \frac{1}{N^* C}\right) + 0
\]

\textbf{Step 2.} For typical parameters $N^* C > 1$, so Tr $> 0$.

\textbf{Step 3.} Compare to standard LV: Tr was \textbf{0} (neutral oscillations). Now Tr $>$ 0.

\textbf{Step 4.} A positive trace means \textbf{at least one eigenvalue has positive real part}. Combined with positive Det, the FP is an \textbf{unstable spiral}.

\medskip\textbf{(d) Why Type II destabilizes:}

The constant-rate $\bar\beta$ assumed predators can consume prey at any density — the more prey, the more they eat, no saturation. Type II adds saturation: with handling time $h_T$, predator consumption levels off at high prey density. When prey become abundant, predators can't keep up, and prey ``escape'' control. This means small perturbations away from the FP grow (oscillations gain amplitude), whereas standard LV had only neutral cycles. In the full nonlinear system, the destabilization typically saturates into a stable limit cycle.

\medskip\textbf{(e) Why no $A$ matrix:}

\textbf{Step 1.} For the $A$ matrix we'd compute $\dot N/N$:
\[
\frac{\dot N}{N} = r - \frac{\alpha P}{1+\alpha h_T N}
\]

\textbf{Step 2.} The coefficient of $P$ here is $-\alpha/(1+\alpha h_T N)$, which depends on $N$. This is \textbf{not} the constant-times-state-variable form $A_{ij}X_j$ required for the interaction matrix.

\textbf{Step 3.} Conclusion: \textbf{cannot} write $A$. The Type II response inherently breaks the bilinear $A_{ij}X_iX_j$ structure that the interaction matrix method requires.
\end{solution}

\subsection{Problem 2.3: Pack Hunting (2024 Q1)}

\begin{problem}
4 predators always hunt in packs to kill 1 prey at a time.
\begin{enumerate}[label=(\alph*)]
\item Choose appropriate compartments.
\item Write the differential equations.
\item Interpret parameters relative to standard LV.
\item Write $A$, find $\text{Tr}$, $\text{Det}$, eigenvalues.
\item Is this system more or less stable?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Compartments:} Let $N$ = prey density, $P_p$ = density of \textbf{packs} (not individual predators). One pack contains 4 predators.

\textbf{(b) Equations:} A pack consumes prey at rate $\bar\beta_p N P_p$. Each kill feeds the pack:
\[
\frac{dN}{dt} = rN - \bar\beta_p N P_p
\]
\[
\frac{dP_p}{dt} = \varepsilon \bar\beta_p N P_p - Z_p P_p
\]
Same structure as standard LV; only the units of $P_p$ and consumption rate change.

\textbf{(c) Parameters:}
\begin{itemize}
\item $r$: same units (prey/time per prey).
\item $\bar\beta_p$: consumption rate per pack. Since each pack eats as much as 4 individual predators in coordinated kills, $\bar\beta_p \approx 4\bar\beta$ when comparing to individual predators.
\item $\varepsilon$: still dimensionless, prey-biomass-to-predator-biomass conversion. Same value.
\item $Z_p$: pack mortality. Hard to define exactly — average pack persistence rate. Roughly same magnitude as $Z$.
\end{itemize}

\textbf{(d) Interaction matrix:}
\[
A = \begin{pmatrix} 0 & -\bar\beta_p \\ \varepsilon\bar\beta_p & 0 \end{pmatrix}
\]
$\text{Tr}(A) = 0$, $\text{Det}(A) = \varepsilon\bar\beta_p^2$. Eigenvalues:
\[
\lambda_\pm = \pm i\bar\beta_p \sqrt{\varepsilon}
\]

\textbf{(e) Stability:} Eigenvalues are pure imaginary $\Rightarrow$ neutral limit cycles (same conclusion as LV). However, since $\bar\beta_p \approx 4\bar\beta$, the magnitude $|\lambda|$ is 4$\times$ larger, meaning \textbf{larger-amplitude oscillations}. The pack system has bigger boom-bust cycles, getting closer to extinction in the troughs. Strictly: same stability class. Loosely: \textbf{less stable in practice} due to extinction risk.
\end{solution}

\subsection{Problem 2.4: Cannibalistic Predator (2021 Q2)}

\begin{problem}
Modify LV for cannibalistic predator:
\[
\frac{dN}{dt} = rN - \bar\beta NP, \quad \frac{dP}{dt} = \varepsilon\bar\beta NP - (1-\varepsilon)\bar\beta_c P^2 - ZP
\]
\begin{enumerate}[label=(\alph*)]
\item Draw the interaction network.
\item Write $A$. Find $\text{Tr}$ and $\text{Det}$.
\item Find fixed points.
\item Compute $J^*$ at non-trivial FP.
\item Find eigenvalues. What happens when $\bar\beta_c \gg \bar\beta$?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Network diagram:} Two nodes (prey $N$, predator $P$). Arrow from $N$ to $P$ representing predation. \textbf{Self-loop on $P$} representing cannibalism (predator eats predator).

\medskip\textbf{(b) Interaction matrix:}

\textbf{Step 1.} Divide each equation by its variable. For $\dot N$:
\[
\frac{1}{N}\frac{dN}{dt} = r - \bar\beta P
\]

\textbf{Step 2.} For $\dot P$:
\[
\frac{1}{P}\frac{dP}{dt} = \varepsilon\bar\beta N - (1-\varepsilon)\bar\beta_c P - Z
\]
\stepwhy{cancel one $P$ from $P^2$ term to get $-(1-\varepsilon)\bar\beta_c P$}

\textbf{Step 3.} Read coefficients:
\begin{itemize}
\item In $r - \bar\beta P$: coeff of $N$ is 0; coeff of $P$ is $-\bar\beta$.
\item In $\varepsilon\bar\beta N - (1-\varepsilon)\bar\beta_c P - Z$: coeff of $N$ is $\varepsilon\bar\beta$; coeff of $P$ is $-(1-\varepsilon)\bar\beta_c$.
\end{itemize}

\textbf{Step 4.}
\[
A = \begin{pmatrix} 0 & -\bar\beta \\ \varepsilon\bar\beta & -(1-\varepsilon)\bar\beta_c \end{pmatrix}
\]

\textbf{Step 5.} Trace: sum of diagonal:
\[
\text{Tr}(A) = 0 + [-(1-\varepsilon)\bar\beta_c] = -(1-\varepsilon)\bar\beta_c
\]
Since $0 < \varepsilon < 1$ and $\bar\beta_c > 0$: \textbf{Tr$(A) < 0$ — stabilizing!}

\textbf{Step 6.} Determinant:
\[
\text{Det}(A) = (0)\cdot[-(1-\varepsilon)\bar\beta_c] - (-\bar\beta)(\varepsilon\bar\beta) = 0 + \varepsilon\bar\beta^2 = \varepsilon\bar\beta^2 > 0
\]

\medskip\textbf{(c) Fixed points:}

\textbf{Step 1.} Set $\dot N = 0$:
\[
N(r - \bar\beta P) = 0 \thus N = 0 \text{ or } P = r/\bar\beta
\]

\textbf{Step 2.} Set $\dot P = 0$:
\[
\varepsilon\bar\beta NP - (1-\varepsilon)\bar\beta_c P^2 - ZP = 0
\]

\textbf{Step 3.} Factor $P$:
\[
P[\varepsilon\bar\beta N - (1-\varepsilon)\bar\beta_c P - Z] = 0
\]

\textbf{Step 4.} Either $P = 0$ or $\varepsilon\bar\beta N - (1-\varepsilon)\bar\beta_c P - Z = 0$. Solve the second:
\[
\varepsilon\bar\beta N = (1-\varepsilon)\bar\beta_c P + Z
\]

\textbf{Step 5.} Solve for $N$:
\[
N = \frac{(1-\varepsilon)\bar\beta_c P + Z}{\varepsilon\bar\beta}
\]

\textbf{Step 6.} Substitute $P^* = r/\bar\beta$:
\[
N^* = \frac{(1-\varepsilon)\bar\beta_c\cdot(r/\bar\beta) + Z}{\varepsilon\bar\beta}
\]

\textbf{Step 7.} Get common denominator in numerator: $(1-\varepsilon)\bar\beta_c r/\bar\beta + Z = [(1-\varepsilon)\bar\beta_c r + Z\bar\beta]/\bar\beta$:
\[
N^* = \frac{(1-\varepsilon)\bar\beta_c r + Z\bar\beta}{\bar\beta\cdot\varepsilon\bar\beta} = \frac{Z\bar\beta + (1-\varepsilon)\bar\beta_c r}{\varepsilon\bar\beta^2}
\]
\stepwhy{multiply both sides by $\bar\beta$, then the $\bar\beta\cdot\varepsilon\bar\beta = \varepsilon\bar\beta^2$ in denominator}
\[
\boxed{N^* = \frac{Z\bar\beta + (1-\varepsilon)\bar\beta_c r}{\varepsilon\bar\beta^2}, \quad P^* = \frac{r}{\bar\beta}}
\]

\medskip\textbf{(d) Standard Jacobian:}

\textbf{Step 1.} Define $f = rN - \bar\beta NP$ and $g = \varepsilon\bar\beta NP - (1-\varepsilon)\bar\beta_c P^2 - ZP$.

\textbf{Step 2.} Partials of $f$ (same as standard LV):
\[
\frac{\partial f}{\partial N} = r - \bar\beta P, \quad \frac{\partial f}{\partial P} = -\bar\beta N
\]

\textbf{Step 3.} Partial of $g$ w.r.t.\ $N$:
\[
\frac{\partial g}{\partial N} = \varepsilon\bar\beta P - 0 - 0 = \varepsilon\bar\beta P
\]

\textbf{Step 4.} Partial of $g$ w.r.t.\ $P$. Note $\partial[P^2]/\partial P = 2P$:
\[
\frac{\partial g}{\partial P} = \varepsilon\bar\beta N - 2(1-\varepsilon)\bar\beta_c P - Z
\]

\textbf{Step 5.} Build Jacobian:
\[
J = \begin{pmatrix} r - \bar\beta P & -\bar\beta N \\ \varepsilon\bar\beta P & \varepsilon\bar\beta N - 2(1-\varepsilon)\bar\beta_c P - Z \end{pmatrix}
\]

\textbf{Step 6.} Evaluate at FP. Entry (1,1):
\[
r - \bar\beta P^* = r - \bar\beta\cdot\frac{r}{\bar\beta} = r - r = 0
\]

\textbf{Step 7.} Entry (1,2): $-\bar\beta N^*$ (keep symbolic).

\textbf{Step 8.} Entry (2,1):
\[
\varepsilon\bar\beta P^* = \varepsilon\bar\beta\cdot\frac{r}{\bar\beta} = \varepsilon r
\]

\textbf{Step 9.} Entry (2,2). Use the FP condition $\varepsilon\bar\beta N^* - (1-\varepsilon)\bar\beta_c P^* - Z = 0$, which gives $\varepsilon\bar\beta N^* - Z = (1-\varepsilon)\bar\beta_c P^*$:
\[
\varepsilon\bar\beta N^* - 2(1-\varepsilon)\bar\beta_c P^* - Z = [(1-\varepsilon)\bar\beta_c P^*] - 2(1-\varepsilon)\bar\beta_c P^*
\]

\textbf{Step 10.} Simplify: $X - 2X = -X$:
\[
= -(1-\varepsilon)\bar\beta_c P^*
\]

\textbf{Step 11.} Substitute $P^* = r/\bar\beta$:
\[
= -(1-\varepsilon)\bar\beta_c\cdot\frac{r}{\bar\beta} = -\frac{(1-\varepsilon)\bar\beta_c r}{\bar\beta}
\]

\textbf{Step 12.} Final $J^*$:
\[
\boxed{J^* = \begin{pmatrix} 0 & -\bar\beta N^* \\ \varepsilon r & -(1-\varepsilon)\bar\beta_c r/\bar\beta \end{pmatrix}}
\]

\medskip\textbf{(e) Eigenvalues:}

\textbf{Step 1.} Trace:
\[
\text{Tr}(J^*) = 0 + \left[-\frac{(1-\varepsilon)\bar\beta_c r}{\bar\beta}\right] = -\frac{(1-\varepsilon)\bar\beta_c r}{\bar\beta} < 0
\]

\textbf{Step 2.} Determinant: $ad - bc$ with $a = 0$, $b = -\bar\beta N^*$, $c = \varepsilon r$, $d = -(1-\varepsilon)\bar\beta_c r/\bar\beta$:
\[
\text{Det}(J^*) = (0)\cdot d - (-\bar\beta N^*)(\varepsilon r) = 0 + \varepsilon r\bar\beta N^* = \varepsilon r\bar\beta N^* > 0
\]

\textbf{Step 3.} Apply formula:
\[
\lambda_\pm = \frac{\text{Tr}}{2}\left(1 \pm \sqrt{1 - \frac{4\text{Det}}{\text{Tr}^2}}\right)
\]

\textbf{Step 4.} Tr $< 0$ and Det $> 0$ guarantee $\text{Re}(\lambda_\pm) < 0$ regardless of discriminant sign $\Rightarrow$ \textbf{stable}.

\textbf{Step 5.} Discriminant $\Delta = 1 - 4\text{Det}/\text{Tr}^2$. Substitute:
\[
\Delta = 1 - \frac{4\varepsilon r\bar\beta N^*}{[(1-\varepsilon)\bar\beta_c r/\bar\beta]^2} = 1 - \frac{4\varepsilon r\bar\beta N^*\bar\beta^2}{(1-\varepsilon)^2\bar\beta_c^2 r^2}
\]

\textbf{Step 6.} Simplify: $4\varepsilon\bar\beta^3 N^*/((1-\varepsilon)^2\bar\beta_c^2 r)$:
\[
\Delta = 1 - \frac{4\varepsilon\bar\beta^3 N^*}{(1-\varepsilon)^2\bar\beta_c^2 r}
\]

\textbf{Step 7.} \textbf{Limit $\bar\beta_c \gg \bar\beta$:} The fraction in $\Delta$ scales as $1/\bar\beta_c^2$, which goes to 0. So $\Delta \to 1 > 0$.

\textbf{Step 8.} Positive discriminant $\Rightarrow$ real eigenvalues:
\[
\lambda_\pm = \frac{\text{Tr}}{2}(1 \pm \sqrt\Delta)
\]
Both negative since Tr $< 0$. \textbf{No oscillations: stable node.}

\textbf{Conclusion:} Strong cannibalism eliminates the boom-bust cycles entirely. Predators self-buffer against prey scarcity by eating each other.
\end{solution}

\subsection{Problem 2.5: Three-Species System (2023 Q2)}

\begin{problem}
Pet ownership system: humans $H$, pets $D$, predators $P$.
\[
\frac{dH}{dt} = rH(1 - H/K) + \alpha_D HD
\]
\[
\frac{dD}{dt} = rD(1-D/K) + \omega\alpha_D HD - \bar\beta DP
\]
\[
\frac{dP}{dt} = \varepsilon\bar\beta DP - ZP
\]
\begin{enumerate}[label=(\alph*)]
\item How does this differ from standard predator-prey?
\item Write the interaction matrix $A$.
\item Find fixed points.
\item Compute Jacobian at coexistence FP.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Differences from standard predator-prey:}

\begin{itemize}
\item \textbf{Three compartments} instead of two ($H$, $D$, $P$).
\item \textbf{Mutualism between $H$ and $D$}: term $+\alpha_D HD$ appears in BOTH $\dot H$ and (with weight $\omega$) $\dot D$. Both benefit from contact.
\item \textbf{Asymmetric weights}: parameter $\omega$ allows different benefits ($H$ vs $D$).
\item \textbf{Logistic growth} in both $H$ and $D$ (carrying capacity $K$ each).
\item Compartments are concrete species (humans, pets, predators), not abstractions.
\end{itemize}

\medskip\textbf{(b) Interaction matrix:}

\textbf{Step 1.} Divide each equation by its variable. For $\dot H$:
\[
\frac{1}{H}\frac{dH}{dt} = r(1-H/K) + \alpha_D D
\]

\textbf{Step 2.} Distribute $r$:
\[
= r - \frac{r H}{K} + \alpha_D D
\]

\textbf{Step 3.} For $\dot D$:
\[
\frac{1}{D}\frac{dD}{dt} = r - \frac{rD}{K} + \omega\alpha_D H - \bar\beta P
\]

\textbf{Step 4.} For $\dot P$:
\[
\frac{1}{P}\frac{dP}{dt} = \varepsilon\bar\beta D - Z
\]

\textbf{Step 5.} Read coefficients of state variables ($H$, $D$, $P$) in each per-capita rate:
\begin{itemize}
\item $\dot H/H$: $H$ coeff $= -r/K$, $D$ coeff $= \alpha_D$, $P$ coeff $= 0$.
\item $\dot D/D$: $H$ coeff $= \omega\alpha_D$, $D$ coeff $= -r/K$, $P$ coeff $= -\bar\beta$.
\item $\dot P/P$: $H$ coeff $= 0$, $D$ coeff $= \varepsilon\bar\beta$, $P$ coeff $= 0$.
\end{itemize}

\textbf{Step 6.}
\[
\boxed{A = \begin{pmatrix} -r/K & \alpha_D & 0 \\ \omega\alpha_D & -r/K & -\bar\beta \\ 0 & \varepsilon\bar\beta & 0 \end{pmatrix}}
\]

\textbf{Step 7.} Trace:
\[
\text{Tr}(A) = -r/K + (-r/K) + 0 = -2r/K < 0 \quad\text{(stabilizing)}
\]

\medskip\textbf{(c) Fixed points:}

\textbf{Step 1.} At a non-trivial FP, set each per-capita rate to zero (since each $X^* \neq 0$). Three equations:
\[
\text{(eq 1)}: r - rH/K + \alpha_D D = 0
\]
\[
\text{(eq 2)}: r - rD/K + \omega\alpha_D H - \bar\beta P = 0
\]
\[
\text{(eq 3)}: \varepsilon\bar\beta D - Z = 0
\]

\textbf{Step 2.} Solve eq 3 for $D$:
\[
\varepsilon\bar\beta D = Z \thus D^* = \frac{Z}{\varepsilon\bar\beta}
\]

\textbf{Step 3.} Substitute into eq 1:
\[
r - rH^*/K + \alpha_D\cdot\frac{Z}{\varepsilon\bar\beta} = 0
\]

\textbf{Step 4.} Solve for $H^*$:
\[
\frac{rH^*}{K} = r + \frac{\alpha_D Z}{\varepsilon\bar\beta}
\]
\[
H^* = K + \frac{K\alpha_D Z}{r\varepsilon\bar\beta} = K\left(1 + \frac{\alpha_D Z}{r\varepsilon\bar\beta}\right)
\]
\stepwhy{multiply both sides by $K/r$, then factor $K$}

\textbf{Step 5.} Substitute $D^*$ and $H^*$ into eq 2 to find $P^*$:
\[
P^* = \frac{r - rD^*/K + \omega\alpha_D H^*}{\bar\beta}
\]
(Symbolic; positive if $\omega\alpha_D H^* > rD^*/K - r$.)

\medskip\textbf{(d) Jacobian at the coexistence FP:}

\textbf{Step 1.} Define $f_1 = rH(1-H/K) + \alpha_D HD$ etc. Compute partials.

\textbf{Step 2.} For $f_1 = rH - rH^2/K + \alpha_D HD$:
\[
\partial f_1/\partial H = r - 2rH/K + \alpha_D D
\]
\[
\partial f_1/\partial D = \alpha_D H, \qquad \partial f_1/\partial P = 0
\]

\textbf{Step 3.} For $f_2 = rD - rD^2/K + \omega\alpha_D HD - \bar\beta DP$:
\[
\partial f_2/\partial H = \omega\alpha_D D
\]
\[
\partial f_2/\partial D = r - 2rD/K + \omega\alpha_D H - \bar\beta P
\]
\[
\partial f_2/\partial P = -\bar\beta D
\]

\textbf{Step 4.} For $f_3 = \varepsilon\bar\beta DP - ZP$:
\[
\partial f_3/\partial H = 0, \quad \partial f_3/\partial D = \varepsilon\bar\beta P, \quad \partial f_3/\partial P = \varepsilon\bar\beta D - Z
\]

\textbf{Step 5.} Build $J$:
\[
J = \begin{pmatrix} r - 2rH/K + \alpha_D D & \alpha_D H & 0 \\ \omega\alpha_D D & r - 2rD/K + \omega\alpha_D H - \bar\beta P & -\bar\beta D \\ 0 & \varepsilon\bar\beta P & \varepsilon\bar\beta D - Z \end{pmatrix}
\]

\textbf{Step 6.} Simplify (1,1) at the FP. Use eq 1 ($r + \alpha_D D^* = rH^*/K$):
\[
r - 2rH^*/K + \alpha_D D^* = (r + \alpha_D D^*) - 2rH^*/K = rH^*/K - 2rH^*/K = -rH^*/K
\]

\textbf{Step 7.} Similarly (2,2): use eq 2 ($r + \omega\alpha_D H^* - \bar\beta P^* = rD^*/K$):
\[
r - 2rD^*/K + \omega\alpha_D H^* - \bar\beta P^* = rD^*/K - 2rD^*/K = -rD^*/K
\]

\textbf{Step 8.} (3,3) entry: from eq 3, $\varepsilon\bar\beta D^* - Z = 0$.

\textbf{Step 9.} Final $J^*$:
\[
J^* = \begin{pmatrix} -rH^*/K & \alpha_D H^* & 0 \\ \omega\alpha_D D^* & -rD^*/K & -\bar\beta D^* \\ 0 & \varepsilon\bar\beta P^* & 0 \end{pmatrix}
\]

\textbf{Step 10.} Trace:
\[
\text{Tr}(J^*) = -rH^*/K - rD^*/K + 0 = -\frac{r(H^* + D^*)}{K} < 0
\]

\textbf{Step 11.} Determinant. Expand along first column ($a_{11} = -rH^*/K$, then 0, 0):
\[
\text{Det}(J^*) = -\frac{rH^*}{K}\det\begin{pmatrix} -rD^*/K & -\bar\beta D^* \\ \varepsilon\bar\beta P^* & 0 \end{pmatrix} - 0 + 0
\]

\textbf{Step 12.} Compute the 2$\times$2 sub-det: $(-rD^*/K)(0) - (-\bar\beta D^*)(\varepsilon\bar\beta P^*) = 0 + \varepsilon\bar\beta^2 D^* P^*$:
\[
\text{Det}(J^*) = -\frac{rH^*}{K}\cdot\varepsilon\bar\beta^2 D^* P^* < 0
\]

\textbf{Step 13.} For 3$\times$3 stability via Routh-Hurwitz: characteristic polynomial $\lambda^3 + a_2\lambda^2 + a_1\lambda + a_0 = 0$ with $a_2 = -\text{Tr}(J^*) > 0$, $a_0 = -\text{Det}(J^*) > 0$, and need $a_2 a_1 > a_0$.

\textbf{Step 14.} Both Tr $< 0$ and Det $< 0$ have correct sign for stability. Combined with sufficient logistic damping (large $r/K$), the third Routh-Hurwitz condition is satisfied. \textbf{Likely stable.}
\end{solution}

\subsection{Problem 2.6: LV with Logistic Prey (2017 Q2)}

\begin{problem}
Modify standard LV by adding logistic self-regulation to the prey:
\[
\frac{dN}{dt} = rN - \frac{r}{K}N^2 - \bar\beta NP, \qquad \frac{dP}{dt} = \varepsilon\bar\beta NP - ZP
\]
\begin{enumerate}[label=(\alph*)]
\item Write the interaction matrix $A$ and explain how it differs from standard LV.
\item Find the non-trivial fixed point $(N^*, P^*)$.
\item Compute $\text{Tr}(A)$ and $\text{Det}(A)$. Are eigenvalues real or complex?
\item Classify stability. Is this more or less stable than standard LV?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Interaction matrix:}

\textbf{Step 1.} Divide $\dot N$ by $N$:
\[
\frac{\dot N}{N} = r - \frac{r}{K}N - \bar\beta P
\]
\stepwhy{cancel one $N$ from $rN^2/K$ to get $rN/K$}

\textbf{Step 2.} Divide $\dot P$ by $P$:
\[
\frac{\dot P}{P} = \varepsilon\bar\beta N - Z
\]

\textbf{Step 3.} Read off coefficients of $N$ and $P$:
\begin{itemize}
\item $\dot N/N$: coeff of $N$ is $-r/K$ (self-interaction), coeff of $P$ is $-\bar\beta$.
\item $\dot P/P$: coeff of $N$ is $\varepsilon\bar\beta$, coeff of $P$ is 0.
\end{itemize}

\textbf{Step 4.}
\[
A = \begin{pmatrix} -r/K & -\bar\beta \\ \varepsilon\bar\beta & 0 \end{pmatrix}
\]

\textbf{Difference from standard LV:} The (1,1) entry is $-r/K < 0$ instead of 0. This is the self-regulation (intraspecific competition) from the logistic term.

\medskip\textbf{(b) Fixed points:}

\textbf{Step 1.} Set per-capita rates to zero:
\[
r - \frac{r}{K}N - \bar\beta P = 0 \quad\text{(prey eq)}
\]
\[
\varepsilon\bar\beta N - Z = 0 \thus N^* = \frac{Z}{\varepsilon\bar\beta}
\]

\textbf{Step 2.} Substitute $N^*$ into prey equation:
\[
r - \frac{r}{K}\cdot\frac{Z}{\varepsilon\bar\beta} - \bar\beta P^* = 0
\]

\textbf{Step 3.} Solve for $P^*$:
\[
\bar\beta P^* = r - \frac{rZ}{K\varepsilon\bar\beta} = r\left(1 - \frac{Z}{K\varepsilon\bar\beta}\right)
\]
\[
P^* = \frac{r}{\bar\beta}\left(1 - \frac{Z}{K\varepsilon\bar\beta}\right) = \frac{r}{\bar\beta} - \frac{rZ}{K\varepsilon\bar\beta^2}
\]
\stepwhy{divide by $\bar\beta$}

\textbf{Step 4.} Biological constraint: $P^* > 0$ requires $1 > Z/(K\varepsilon\bar\beta)$, i.e., prey carrying capacity large enough to sustain predators.

\medskip\textbf{(c) Tr, Det, discriminant:}

\textbf{Step 1.} Trace:
\[
\text{Tr}(A) = -r/K + 0 = -r/K < 0
\]

\textbf{Step 2.} Determinant:
\[
\text{Det}(A) = (-r/K)(0) - (-\bar\beta)(\varepsilon\bar\beta) = 0 + \varepsilon\bar\beta^2 = \varepsilon\bar\beta^2 > 0
\]

\textbf{Step 3.} Discriminant $\Delta = \text{Tr}^2 - 4\text{Det}$:
\[
\Delta = \frac{r^2}{K^2} - 4\varepsilon\bar\beta^2
\]

If $\Delta < 0$ (i.e., $4\varepsilon\bar\beta^2 > r^2/K^2$, or $K > r/(2\bar\beta\sqrt\varepsilon)$): \textbf{complex eigenvalues (spiral)}.\\
If $\Delta > 0$: \textbf{real eigenvalues (node)}.

\medskip\textbf{(d) Stability classification:}

\textbf{Step 1.} Tr $< 0$ and Det $> 0$ $\Rightarrow$ \textbf{stable} in all cases.

\textbf{Step 2.} If $\Delta < 0$ (common when coupling is strong): \textbf{stable spiral} — oscillations that dampen into the equilibrium.

\textbf{Step 3.} If $\Delta > 0$: \textbf{stable node} — smooth exponential decay to equilibrium, no oscillations.

\textbf{Step 4.} Comparison with standard LV (Tr $= 0$, neutral oscillations):
\begin{itemize}
\item Adding logistic self-regulation shifts Tr from 0 to $-r/K < 0$.
\item This converts neutral (undamped) oscillations into \textbf{damped} oscillations.
\item The system is \textbf{more stable}: perturbations decay rather than persist forever.
\end{itemize}
\end{solution}

\subsection{Problem 2.7: Prion Disease as LV (2018 Q2)}

\begin{problem}
Prion disease: normal protein $Y_1$ converts to misfolded $Y_2$ upon contact. Treat analogously to predator-prey:
\[
\frac{dY_1}{dt} = rY_1 - \bar\beta Y_1 Y_2, \qquad \frac{dY_2}{dt} = \varepsilon\bar\beta Y_1 Y_2 - Z Y_2
\]
\begin{enumerate}[label=(\alph*)]
\item Interpret each parameter in the prion context.
\item Which parameters might have different values than in a biological predator-prey system? Why?
\item Find fixed points and compute eigenvalues at the coexistence FP.
\item What would happen if $r = 0$ (no natural production of $Y_1$)? Interpret biologically.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Parameter interpretation:}

\begin{itemize}
\item $r$: production rate of normal protein $Y_1$ (e.g., by ribosomes). Unlike prey, $Y_1$ doesn't self-replicate by interaction — it's produced continuously.
\item $\bar\beta$: conversion rate when $Y_1$ contacts $Y_2$ (misfolding induction rate). Dimensionless-ish: rate per $Y_1$ per $Y_2$ per time.
\item $\varepsilon$: conversion efficiency. Each $Y_1$ consumed creates $\varepsilon$ units of $Y_2$. In prion context $\varepsilon = 1$ exactly (every normal protein that misfolds becomes one prion), unlike biological predation where $\varepsilon < 1$.
\item $Z$: degradation rate of misfolded protein $Y_2$ (by proteasome, chaperones, etc.).
\end{itemize}

\medskip\textbf{(b) Parameters with different values:}

\textbf{Step 1.} $\varepsilon$: Likely $\varepsilon = 1$ for prions (exact 1-to-1 conversion of $Y_1$ into $Y_2$). In predator-prey, $\varepsilon \ll 1$ (inefficient energy conversion).

\textbf{Step 2.} $r$: In predator-prey, prey reproduce by breeding (proportional to $N$). Here $Y_1$ is produced at a constant rate from cellular machinery — so really $r$ is not a per-capita rate but an absolute production rate. The equation should perhaps be $\dot Y_1 = r_0 - \bar\beta Y_1 Y_2$ with constant $r_0$. Using $rY_1$ approximates this when $Y_1$ stays near a baseline.

\textbf{Step 3.} $Z$: Degradation of $Y_2$ is very slow in prion diseases (that's why they're lethal) — $Z$ is much smaller than in typical predator-prey.

\medskip\textbf{(c) Fixed points and eigenvalues:}

\textbf{Step 1.} Same algebra as standard LV. Non-trivial FP:
\[
N^* = Y_1^* = Z/(\varepsilon\bar\beta), \quad P^* = Y_2^* = r/\bar\beta
\]

\textbf{Step 2.} Interaction matrix:
\[
A = \begin{pmatrix} 0 & -\bar\beta \\ \varepsilon\bar\beta & 0 \end{pmatrix}
\]

\textbf{Step 3.} Tr $= 0$, Det $= \varepsilon\bar\beta^2$. Eigenvalues:
\[
\lambda_\pm = \pm i\bar\beta\sqrt\varepsilon
\]
\textbf{Neutral oscillations.} In the prion context: normal and misfolded protein oscillate indefinitely.

\medskip\textbf{(d) What if $r = 0$?}

\textbf{Step 1.} With $r = 0$: $\dot Y_1 = -\bar\beta Y_1 Y_2$. The fixed point $Y_2^* = r/\bar\beta = 0$ collapses.

\textbf{Step 2.} Equations become:
\[
\dot Y_1 = -\bar\beta Y_1 Y_2, \quad \dot Y_2 = \varepsilon\bar\beta Y_1 Y_2 - Z Y_2
\]

\textbf{Step 3.} Without $Y_1$ production, the only steady state is $Y_1 = 0$ or $Y_2 = 0$.

\textbf{Step 4.} Biological interpretation: if the cell stops producing normal protein $Y_1$ (e.g., gene silenced, disease state) then $Y_1$ will be gradually consumed until it is depleted. Then $Y_2$ will degrade at rate $Z$ (no more substrate). Long-term: both go to zero. \textbf{The disease could self-limit!} This is actually the basis of some prion therapy approaches: reduce $Y_1$ production to starve $Y_2$.
\end{solution}

\subsection{Problem 2.8: Competing Species}

\begin{problem}
Two species $N_1$ and $N_2$ compete for the same resource. Each grows logistically and is harmed by the other:
\[
\frac{dN_1}{dt} = r_1 N_1\left(1 - \frac{N_1}{K_1} - \frac{\alpha_{12}}{K_1}N_2\right)
\]
\[
\frac{dN_2}{dt} = r_2 N_2\left(1 - \frac{N_2}{K_2} - \frac{\alpha_{21}}{K_2}N_1\right)
\]
\begin{enumerate}[label=(\alph*)]
\item Write the interaction matrix $A$.
\item Find all four fixed points.
\item Under what conditions does $N_1$ win? $N_2$ win? Coexistence? Bistability?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Interaction matrix:}

\textbf{Step 1.} Divide $\dot N_1$ by $N_1$:
\[
\frac{\dot N_1}{N_1} = r_1\left(1 - \frac{N_1}{K_1} - \frac{\alpha_{12}N_2}{K_1}\right) = r_1 - \frac{r_1}{K_1}N_1 - \frac{r_1\alpha_{12}}{K_1}N_2
\]

\textbf{Step 2.} Divide $\dot N_2$ by $N_2$ similarly.

\textbf{Step 3.}
\[
A = \begin{pmatrix} -r_1/K_1 & -r_1\alpha_{12}/K_1 \\ -r_2\alpha_{21}/K_2 & -r_2/K_2 \end{pmatrix}
\]

\medskip\textbf{(b) Four fixed points:}

\textbf{FP 1:} Extinction of both: $(N_1^*, N_2^*) = (0, 0)$.

\textbf{FP 2:} Only $N_1$ survives ($N_2 = 0$): set $\dot N_1/N_1 = 0$ with $N_2 = 0$: $r_1(1 - N_1/K_1) = 0 \Rightarrow N_1^* = K_1$.
\[
(K_1, 0)
\]

\textbf{FP 3:} Only $N_2$ survives ($N_1 = 0$): similarly $(0, K_2)$.

\textbf{FP 4:} Coexistence. Solve simultaneously:
\[
1 - N_1/K_1 - \alpha_{12}N_2/K_1 = 0
\]
\[
1 - N_2/K_2 - \alpha_{21}N_1/K_2 = 0
\]
From eq 1: $N_1 = K_1 - \alpha_{12}N_2$.
Substitute into eq 2:
\[
1 - \frac{N_2}{K_2} - \frac{\alpha_{21}(K_1 - \alpha_{12}N_2)}{K_2} = 0
\]
\[
K_2 - N_2 - \alpha_{21}K_1 + \alpha_{21}\alpha_{12}N_2 = 0
\]
\[
N_2(1 - \alpha_{21}\alpha_{12}) = K_2 - \alpha_{21}K_1
\]
\[
N_2^* = \frac{K_2 - \alpha_{21}K_1}{1 - \alpha_{12}\alpha_{21}}, \qquad N_1^* = \frac{K_1 - \alpha_{12}K_2}{1 - \alpha_{12}\alpha_{21}}
\]

\medskip\textbf{(c) Competition outcomes:}

The key quantity is the product $\alpha_{12}\alpha_{21}$ (mutual inhibition strength).

\textbf{$N_1$ wins (competitive exclusion):} $K_1 > \alpha_{12}K_2$ and $K_2 < \alpha_{21}K_1$.
\stepwhy{$N_1$ can grow when $N_2$ is at its carrying capacity, but not vice versa}

\textbf{$N_2$ wins (competitive exclusion):} $K_2 > \alpha_{21}K_1$ and $K_1 < \alpha_{12}K_2$.

\textbf{Stable coexistence:} $\alpha_{12}\alpha_{21} < 1$ and $N_1^* > 0$, $N_2^* > 0$.
\stepwhy{Intraspecific competition stronger than interspecific; species limit themselves more than each other}

\textbf{Bistability (unstable coexistence):} $\alpha_{12}\alpha_{21} > 1$.
\stepwhy{Interspecific competition stronger; whichever species is more abundant tends to exclude the other; outcome depends on initial conditions}

\textbf{Biological intuition:} Competitive exclusion (the principle that two species competing for the same resource cannot stably coexist) is recovered when $\alpha_{12}\alpha_{21} \geq 1$. Coexistence is possible only when each species inhibits itself more than it inhibits the competitor.
\end{solution}

\newpage
%==================================================================
\section{Disease (SIR / SIRS) Models}
%==================================================================

\subsection{Problem 3.1: Standard SIR (Foundation)}

\begin{problem}
\[
\frac{dS}{dt} = -\beta SI, \quad \frac{dI}{dt} = \beta SI - \gamma I, \quad \frac{dR}{dt} = \gamma I
\]
\begin{enumerate}[label=(\alph*)]
\item Show $S+I+R$ is conserved.
\item Find all fixed points.
\item Compute $J^*$ at the disease-free state $(S^*=N, I^*=0, R^*=0)$.
\item Find eigenvalues, define $R_0$.
\item When does the disease spread?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Conservation:}

\textbf{Step 1.} Sum all three equations:
\[
\frac{dS}{dt} + \frac{dI}{dt} + \frac{dR}{dt} = (-\beta SI) + (\beta SI - \gamma I) + (\gamma I)
\]

\textbf{Step 2.} Distribute parentheses:
\[
= -\beta SI + \beta SI - \gamma I + \gamma I
\]

\textbf{Step 3.} Cancel pairs:
\[
= 0 + 0 = 0 \checkmark
\]
\stepwhy{$-\beta SI + \beta SI = 0$ and $-\gamma I + \gamma I = 0$}

\textbf{Step 4.} So $\frac{d}{dt}(S+I+R) = 0$, meaning $S+I+R = N$ constant. Eliminate $R = N - S - I$.

\medskip\textbf{(b) Fixed points:}

\textbf{Step 1.} Set $\dot S = 0$:
\[
-\beta SI = 0 \thus S = 0 \text{ or } I = 0
\]

\textbf{Step 2.} Set $\dot I = 0$:
\[
\beta SI - \gamma I = 0
\]

\textbf{Step 3.} Factor $I$:
\[
I(\beta S - \gamma) = 0 \thus I = 0 \text{ or } S = \gamma/\beta
\]
\stepwhy{solving $\beta S - \gamma = 0$ for $S$}

\textbf{Step 4.} Combine: \textbf{disease-free state} $I^* = 0$, $S^* = N$ (free; together with conservation $R^* = 0$).

\textbf{Step 5.} \textbf{Endemic possibility:} $S^* = \gamma/\beta$, $I^* > 0$. But this requires $\gamma/\beta < N$, i.e., $R_0 = \beta N/\gamma > 1$.

\medskip\textbf{(c) Jacobian at disease-free FP:}

Define $f(S,I) = -\beta SI$, $g(S,I) = \beta SI - \gamma I$. (Eliminated $R$ via conservation.)

\textbf{Step 1.} Partials of $f$:
\[
\frac{\partial f}{\partial S} = \frac{\partial}{\partial S}[-\beta SI] = -\beta I
\]
\stepwhy{$I$ is constant w.r.t.\ $S$}

\textbf{Step 2.}
\[
\frac{\partial f}{\partial I} = -\beta S
\]

\textbf{Step 3.} Partials of $g$:
\[
\frac{\partial g}{\partial S} = \beta I
\]

\textbf{Step 4.}
\[
\frac{\partial g}{\partial I} = \beta S - \gamma
\]

\textbf{Step 5.} Build $J$:
\[
J = \begin{pmatrix} -\beta I & -\beta S \\ \beta I & \beta S - \gamma \end{pmatrix}
\]

\textbf{Step 6.} Evaluate at $(S^*, I^*) = (N, 0)$. Plug in $I = 0$ and $S = N$:
\[
J^* = \begin{pmatrix} -\beta\cdot 0 & -\beta\cdot N \\ \beta\cdot 0 & \beta N - \gamma \end{pmatrix} = \begin{pmatrix} 0 & -\beta N \\ 0 & \beta N - \gamma \end{pmatrix}
\]
\stepwhy{$\beta\cdot 0 = 0$}

\medskip\textbf{(d) Eigenvalues:}

\textbf{Step 1.} Notice $J^*$ is upper triangular (the (2,1) entry is 0).

\textbf{Step 2.} For triangular matrices, eigenvalues are the diagonal entries:
\[
\lambda_1 = 0, \qquad \lambda_2 = \beta N - \gamma
\]

\textbf{Step 3.} Factor $\gamma$ from $\lambda_2$:
\[
\lambda_2 = \beta N - \gamma = \gamma\left(\frac{\beta N}{\gamma} - 1\right) = \gamma(R_0 - 1)
\]
where
\[
\boxed{R_0 = \frac{\beta N}{\gamma}}
\]

\medskip\textbf{(e) Stability conditions:}

\begin{itemize}
\item $R_0 > 1$: $\lambda_2 = \gamma(R_0-1) > 0$, disease-free FP \textbf{unstable}, disease invades.
\item $R_0 < 1$: $\lambda_2 < 0$, disease-free FP \textbf{stable}, disease dies out.
\item $\lambda_1 = 0$: marginal direction along the continuous family of disease-free FPs (susceptible-direction freedom).
\end{itemize}

\textbf{Why the $A$ matrix fails here:} $I^* = 0$ at the disease-free FP. The interaction matrix method divides by state variables ($\dot I/I$), which is undefined when $I = 0$. Must use standard Jacobian.
\end{solution}

\subsection{Problem 3.2: SIRS with Loss of Immunity (2024 Q3 style)}

\begin{problem}
\[
\frac{dS}{dt} = -\beta SI + \nu R, \quad \frac{dI}{dt} = \beta SI - \gamma I, \quad \frac{dR}{dt} = \gamma I - \nu R
\]
\begin{enumerate}[label=(\alph*)]
\item Apply QSSA-like reasoning: which compartment to set $\dot{}$ to zero?
\item Set $dS/dt = 0$ and solve for $dI/dt$ as function of $S$.
\item Try other QSSA choices and compare.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Which compartment to zero out:}

\textbf{Step 1.} Recall MM: $S + E \rightleftharpoons [SE] \to P + E$. We set $d[SE]/dt = 0$ because the complex is the fast intermediate.

\textbf{Step 2.} In SIRS, the analog of $[SE]$ (intermediate compartment) is $I$ (between $S$ and $R$). \emph{Naively}, set $dI/dt = 0$.

\textbf{Step 3.} \textbf{But}: $I$ is what we want to predict (analog of product $P$, not complex $[SE]$).

\textbf{Step 4.} Better choice: set $dS/dt = 0$. This treats $S$ as quasi-conserved by quick replenishment from $R$ (rate $\nu$).

\medskip\textbf{(b) Set $dS/dt = 0$:}

\textbf{Step 1.} The condition gives:
\[
-\beta SI + \nu R = 0
\]

\textbf{Step 2.} Solve for $R$:
\[
\nu R = \beta SI \thus R = \frac{\beta SI}{\nu}
\]
\stepwhy{add $\beta SI$ to both sides, divide by $\nu$}

\textbf{Step 3.} Substitute into conservation $N = S + I + R$:
\[
N = S + I + \frac{\beta SI}{\nu}
\]

\textbf{Step 4.} Rearrange to isolate $I$:
\[
N - S = I + \frac{\beta SI}{\nu}
\]

\textbf{Step 5.} Factor $I$ from RHS:
\[
N - S = I\left(1 + \frac{\beta S}{\nu}\right)
\]

\textbf{Step 6.} Get common denominator inside parens:
\[
1 + \frac{\beta S}{\nu} = \frac{\nu + \beta S}{\nu}
\]

\textbf{Step 7.}
\[
N - S = I\cdot\frac{\nu + \beta S}{\nu} \thus I = \frac{(N-S)\nu}{\nu + \beta S}
\]
\stepwhy{multiply both sides by $\nu/(\nu+\beta S)$}

\textbf{Step 8.} Now compute $dI/dt$ from the original equation:
\[
\frac{dI}{dt} = \beta SI - \gamma I = I(\beta S - \gamma)
\]

\textbf{Step 9.} Substitute $I$ from step 7:
\[
\frac{dI}{dt} = \frac{(N-S)\nu}{\nu + \beta S}(\beta S - \gamma) = \frac{(N-S)\nu(\beta S - \gamma)}{\nu + \beta S}
\]

\medskip\textbf{Behavior of $dI/dt$ vs $S$:}
\begin{itemize}
\item $S = N$: numerator factor $(N-S) = 0$, so $\dot I = 0$. (Disease can't grow if everyone is susceptible at $t=0$? No --- this IC point is also a fixed point.)
\item $S = \gamma/\beta$: numerator factor $(\beta S - \gamma) = 0$, so $\dot I = 0$ at the threshold $R_0 = 1$.
\item $S \in (\gamma/\beta, N)$: both numerator factors positive, $\dot I > 0$, disease grows.
\item $S < \gamma/\beta$: $(\beta S - \gamma) < 0$, $\dot I < 0$, disease shrinks.
\end{itemize}

\medskip\textbf{(c) Try alternative: set $dR/dt = 0$:}

\textbf{Step 1.} The condition gives:
\[
\gamma I - \nu R = 0 \thus \nu R = \gamma I \thus R = \gamma I/\nu
\]

\textbf{Step 2.} Substitute into conservation:
\[
N = S + I + \frac{\gamma I}{\nu} = S + I\left(1 + \frac{\gamma}{\nu}\right)
\]

\textbf{Step 3.} Common denominator:
\[
N - S = I\cdot\frac{\nu + \gamma}{\nu}
\]

\textbf{Step 4.} Solve for $I$:
\[
I = \frac{(N-S)\nu}{\nu + \gamma}
\]

\textbf{Step 5.} Substitute into $\dot I = I(\beta S - \gamma)$:
\[
\frac{dI}{dt} = \frac{(N-S)\nu(\beta S - \gamma)}{\nu + \gamma}
\]

\textbf{Step 6.} Same zeros (at $S = N$ and $S = \gamma/\beta$), but the denominator is $\nu + \gamma$ (a constant) instead of $\nu + \beta S$ (depends on $S$). So this version is \textbf{linear in $S$ between zeros}, while the previous was \textbf{hyperbolic}.

\textbf{Conclusion:} Different QSSA choices give different functional forms (linear vs.\ hyperbolic), but the threshold conditions ($R_0 = 1$ at $S = \gamma/\beta$) are the same.
\end{solution}

\subsection{Problem 3.3: SIR with Asymptomatic and Symptomatic (2020 Q4)}

\begin{problem}
Compartments: $S$, $I_a$ (asymptomatic), $I_s$ (symptomatic), $R$. Fraction $f$ become asymptomatic.
\[
\frac{dS}{dt} = -\beta_a I_a S - \beta_s I_s S
\]
\[
\frac{dI_a}{dt} = f\beta_a I_a S + f\beta_s I_s S - \gamma_a I_a
\]
\[
\frac{dI_s}{dt} = (1-f)\beta_a I_a S + (1-f)\beta_s I_s S - \gamma_s I_s
\]
\[
\frac{dR}{dt} = \gamma_a I_a + \gamma_s I_s
\]
\begin{enumerate}[label=(\alph*)]
\item Show conservation. Reduce equations.
\item Find fixed points.
\item Compute Jacobian at disease-free FP.
\item Find eigenvalues for special case $f = 1$ (all asymptomatic).
\item Same for $\beta_a = \beta_s$, $\gamma_a = \gamma_s$.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Conservation:}

\textbf{Step 1.} Sum all four equations:
\[
\dot S + \dot{I_a} + \dot{I_s} + \dot R = (-\beta_a I_a S - \beta_s I_s S) + [f\beta_a I_a S + f\beta_s I_s S - \gamma_a I_a]
\]
\[
\quad + [(1-f)\beta_a I_a S + (1-f)\beta_s I_s S - \gamma_s I_s] + (\gamma_a I_a + \gamma_s I_s)
\]

\textbf{Step 2.} Group terms involving $\beta_a I_a S$:
\[
-\beta_a I_a S + f\beta_a I_a S + (1-f)\beta_a I_a S = \beta_a I_a S(-1 + f + (1-f)) = \beta_a I_a S(0) = 0
\]
\stepwhy{$f + (1-f) = 1$, then $-1 + 1 = 0$}

\textbf{Step 3.} Similarly for $\beta_s I_s S$ terms: sum to 0.

\textbf{Step 4.} The $\gamma_a I_a$ and $\gamma_s I_s$ terms also cancel between $\dot{I_a}, \dot{I_s}$ and $\dot R$.

\textbf{Step 5.} So $\sum \dot X_i = 0$ and $N = S + I_a + I_s + R$ is conserved. Eliminate $R$.

\medskip\textbf{(b) Disease-free FP:}
\[
I_a^* = I_s^* = 0, \quad S^* = N, \quad R^* = 0
\]

\medskip\textbf{(c) Jacobian at disease-free FP (3 variables: $S, I_a, I_s$):}

\textbf{Step 1.} Compute partials. For $f = -\beta_a I_a S - \beta_s I_s S$:
\[
\partial f/\partial S = -\beta_a I_a - \beta_s I_s, \quad \partial f/\partial I_a = -\beta_a S, \quad \partial f/\partial I_s = -\beta_s S
\]

\textbf{Step 2.} For $g = f\beta_a I_a S + f\beta_s I_s S - \gamma_a I_a$:
\[
\partial g/\partial S = f(\beta_a I_a + \beta_s I_s)
\]
\[
\partial g/\partial I_a = f\beta_a S - \gamma_a, \quad \partial g/\partial I_s = f\beta_s S
\]

\textbf{Step 3.} For $h = (1-f)\beta_a I_a S + (1-f)\beta_s I_s S - \gamma_s I_s$:
\[
\partial h/\partial S = (1-f)(\beta_a I_a + \beta_s I_s)
\]
\[
\partial h/\partial I_a = (1-f)\beta_a S, \quad \partial h/\partial I_s = (1-f)\beta_s S - \gamma_s
\]

\textbf{Step 4.} Build $J$ then evaluate at $(N, 0, 0)$. With $I_a = I_s = 0$, the first column becomes all zeros and $S = N$:
\[
J^* = \begin{pmatrix} 0 & -\beta_a N & -\beta_s N \\ 0 & f\beta_a N - \gamma_a & f\beta_s N \\ 0 & (1-f)\beta_a N & (1-f)\beta_s N - \gamma_s \end{pmatrix}
\]

\medskip\textbf{Eigenvalues (general):}

\textbf{Step 1.} First column zero $\Rightarrow \lambda_1 = 0$.

\textbf{Step 2.} Other eigenvalues come from the lower-right 2$\times$2 block:
\[
M = \begin{pmatrix} f\beta_a N - \gamma_a & f\beta_s N \\ (1-f)\beta_a N & (1-f)\beta_s N - \gamma_s \end{pmatrix}
\]

\textbf{Step 3.} Trace:
\[
\text{Tr}(M) = (f\beta_a N - \gamma_a) + ((1-f)\beta_s N - \gamma_s) = f\beta_a N + (1-f)\beta_s N - \gamma_a - \gamma_s
\]

\textbf{Step 4.} Determinant: $ad - bc$:
\[
\text{Det}(M) = (f\beta_a N - \gamma_a)\cdot((1-f)\beta_s N - \gamma_s) - (f\beta_s N)\cdot((1-f)\beta_a N)
\]

\textbf{Step 5.} Expand the first product:
\[
(f\beta_a N)((1-f)\beta_s N) + (f\beta_a N)(-\gamma_s) + (-\gamma_a)((1-f)\beta_s N) + (-\gamma_a)(-\gamma_s)
\]
\[
= f(1-f)\beta_a\beta_s N^2 - f\beta_a N\gamma_s - (1-f)\beta_s N\gamma_a + \gamma_a\gamma_s
\]

\textbf{Step 6.} Subtract second product:
\[
\text{Det}(M) = f(1-f)\beta_a\beta_s N^2 - f\beta_a N\gamma_s - (1-f)\beta_s N\gamma_a + \gamma_a\gamma_s - f(1-f)\beta_a\beta_s N^2
\]

\textbf{Step 7.} The $\beta_a\beta_s N^2$ terms cancel:
\[
\text{Det}(M) = -f\beta_a N\gamma_s - (1-f)\beta_s N\gamma_a + \gamma_a\gamma_s
\]

\textbf{Step 8.} Factor $\gamma_a\gamma_s$:
\[
\text{Det}(M) = \gamma_a\gamma_s\left[1 - \frac{f\beta_a N}{\gamma_a} - \frac{(1-f)\beta_s N}{\gamma_s}\right]
\]
\stepwhy{divide each term by $\gamma_a\gamma_s$ inside}

\medskip\textbf{(d) Case $f = 1$ (all asymptomatic):}

Plug $f = 1$: every coefficient with $(1-f)$ becomes 0:
\[
J^* = \begin{pmatrix} 0 & -\beta_a N & -\beta_s N \\ 0 & \beta_a N - \gamma_a & \beta_s N \\ 0 & 0 & -\gamma_s \end{pmatrix}
\]

\textbf{Step 1.} Triangular (with first column and (3,1), (3,2) zero). Eigenvalues are diagonal:
\[
\lambda_1 = 0, \quad \lambda_2 = \beta_a N - \gamma_a, \quad \lambda_3 = -\gamma_s
\]

\textbf{Step 2.} $\lambda_3 = -\gamma_s < 0$ (always). $\lambda_2 > 0$ iff $\beta_a N > \gamma_a$, i.e.\ $R_0 = \beta_a N/\gamma_a > 1$.

\textbf{Biological meaning:} With $f = 1$, all infections produce asymptomatic. The $I_s$ compartment is unused; only the asymptomatic-disease $R_0$ matters.

\medskip\textbf{(e) Case $\beta_a = \beta_s = \beta$, $\gamma_a = \gamma_s = \gamma$:}

\textbf{Step 1.} Plug in:
\[
M = \begin{pmatrix} f\beta N - \gamma & f\beta N \\ (1-f)\beta N & (1-f)\beta N - \gamma \end{pmatrix}
\]

\textbf{Step 2.} Trace:
\[
\text{Tr}(M) = f\beta N - \gamma + (1-f)\beta N - \gamma = \beta N(f + (1-f)) - 2\gamma = \beta N - 2\gamma
\]

\textbf{Step 3.} Determinant. From step 7 of general case with these substitutions:
\[
\text{Det}(M) = \gamma^2[1 - f\beta N/\gamma - (1-f)\beta N/\gamma] = \gamma^2[1 - \beta N/\gamma] = \gamma(\gamma - \beta N) = -\gamma(\beta N - \gamma)
\]

\textbf{Step 4.} For eigenvalues, factor: $\lambda^2 - (\beta N - 2\gamma)\lambda + (-\gamma)(\beta N - \gamma) = 0$. Try $\lambda_+ = \beta N - \gamma$:
\[
(\beta N - \gamma)^2 - (\beta N - 2\gamma)(\beta N - \gamma) + (-\gamma)(\beta N - \gamma)
\]
Factor $(\beta N - \gamma)$:
\[
= (\beta N - \gamma)[(\beta N - \gamma) - (\beta N - 2\gamma) - \gamma] = (\beta N - \gamma)[\beta N - \gamma - \beta N + 2\gamma - \gamma] = (\beta N - \gamma)\cdot 0 = 0 \checkmark
\]

\textbf{Step 5.} So $\lambda_+ = \beta N - \gamma$. Then $\lambda_- = \text{Tr}(M) - \lambda_+ = (\beta N - 2\gamma) - (\beta N - \gamma) = -\gamma$.

\textbf{Step 6.} Combined eigenvalues of $J^*$: $\lambda = 0,\, \beta N - \gamma,\, -\gamma$.

\textbf{Step 7.} Disease spreads iff $\lambda_+ = \beta N - \gamma > 0$, i.e.\ $R_0 = \beta N/\gamma > 1$. Same as standard SIR.

\textbf{Biological meaning:} When asymptomatic and symptomatic infections are identical (same $\beta$, same $\gamma$), the split into compartments $I_a$ and $I_s$ doesn't change the $R_0$. This makes sense: same biology, just labeled differently.
\end{solution}

\subsection{Problem 3.4: Two-Variant Disease (2021 Q3-Q4)}

\begin{problem}
First virus then variant. Variant has half transmissibility for those who already had the first virus.
\[
\frac{dS}{dt} = -\beta SI - \beta_2 S I_2
\]
\[
\frac{dI}{dt} = \beta SI - \gamma I
\]
\[
\frac{dS_2}{dt} = \gamma I - \frac{\beta_2}{2} S_2 I_2
\]
\[
\frac{dI_2}{dt} = \beta_2 I_2(S + S_2/2) - \gamma I_2
\]
\[
\frac{dR}{dt} = \gamma I_2
\]
\begin{enumerate}[label=(\alph*)]
\item Conservation? Reduce.
\item Find fixed points.
\item Which FP for stability analysis of variant?
\item Compute Jacobian at that FP.
\item Find eigenvalues. Express conditions in terms of $R_0$.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Sum all 5:} cancels to 0. So $N = S + I + S_2 + I_2 + R$ const. Eliminate $R$.

\textbf{(b) Fixed points:} Need $\dot I = 0 \Rightarrow I^* = 0$ or $S^* = \gamma/\beta$. Need $\dot{I_2} = 0 \Rightarrow I_2^* = 0$ or $\beta_2(S^* + S_2^*/2) = \gamma$. The non-trivial coexistence with both viruses circulating doesn't have a clean fixed point in general — must explore special cases:
\begin{itemize}
\item All recovered: $S=I=S_2=I_2=0$, $R=N$.
\item Pre-outbreak: $S=N$, all others 0.
\item \textbf{First virus done, variant not yet}: $I^* = I_2^* = 0$, $S^* = N - S_2^*$ (anyone left over) or $S_2^*$ chosen freely. This is the relevant FP.
\end{itemize}

\textbf{(c)} To assess if variant takes off: perturb around the post-first-virus state, $I^* = I_2^* = R^* = 0$, $S_2^*$ = whatever fraction was infected by first virus, $S^* = N - S_2^*$. Use \textbf{standard Jacobian} (compartments are zero, so $A$-matrix approach fails).

\textbf{(d) Jacobian at $(S, I, S_2, I_2) = (N - S_2^*, 0, S_2^*, 0)$:}
$\partial(\dot S)/\partial S = -\beta I - \beta_2 I_2 \to 0$\\
$\partial(\dot S)/\partial I = -\beta S \to -\beta(N - S_2^*)$\\
$\partial(\dot S)/\partial S_2 = 0$\\
$\partial(\dot S)/\partial I_2 = -\beta_2 S \to -\beta_2(N - S_2^*)$\\
$\partial(\dot I)/\partial I = \beta S - \gamma \to \beta(N-S_2^*) - \gamma$\\
$\partial(\dot{S_2})/\partial I = \gamma$, $\partial(\dot{S_2})/\partial I_2 = -\beta_2 S_2^*/2$\\
$\partial(\dot{I_2})/\partial I_2 = \beta_2(S + S_2/2) - \gamma \to \beta_2(N - S_2^*/2) - \gamma$
\[
J^* = \begin{pmatrix} 0 & -\beta(N-S_2^*) & 0 & -\beta_2(N-S_2^*) \\ 0 & \beta(N-S_2^*)-\gamma & 0 & 0 \\ 0 & \gamma & 0 & -\beta_2 S_2^*/2 \\ 0 & 0 & 0 & \beta_2(N-S_2^*/2)-\gamma \end{pmatrix}
\]

\textbf{(e) Eigenvalues:} Block-triangular with two zero columns ($S$ and $S_2$).
$\lambda_1 = \lambda_2 = 0$ (susceptible directions; these are redundant constraint directions).
$\lambda_3 = \beta(N-S_2^*) - \gamma$ (first virus mode).
$\lambda_4 = \beta_2(N-S_2^*/2) - \gamma$ (variant mode).

\textbf{Conditions:}
\begin{itemize}
\item First virus spreads if $\lambda_3 > 0 \iff \beta(N-S_2^*)/\gamma > 1 \iff R_{0,1} > 1$.
\item Variant spreads if $\lambda_4 > 0 \iff \beta_2(N-S_2^*/2)/\gamma > 1 \iff R_{0,2} > 1$ where $R_{0,2} = \beta_2(N - S_2^*/2)/\gamma$.
\end{itemize}

\textbf{Two separate $R_0$ values} make sense: each strain has its own reproductive number conditional on the susceptible pool (which differs since variant gets reduced spread from $S_2$ population).
\end{solution}

\subsection{Problem 3.5: Vaccination (2021 Q4d-e)}

\begin{problem}
Fraction $f$ vaccinated. Vaccine reduces infection rate $\beta_2 \to \beta_2(1-p_1)$ and increases recovery $\gamma \to \gamma(1+p_2)$ for the vaccinated. Find: how do infection terms change? What's condition for variant to spread when $f = 1$?
\end{problem}

\begin{solution}
For each contact event, both partners might be vaccinated (probability $f^2$), one of two scenarios for one being vaccinated ($2f(1-f)$), or neither ($(1-f)^2$). Effective infection rate (averaged over pairings):
\[
\beta_2^{\text{eff}} = \beta_2[(1-f)^2 + 2f(1-f)(1-p_1) + f^2(1-p_1)^2] = \beta_2[1 - f(1-(1-p_1))]^2 = \beta_2(1-fp_1)^2
\]
Wait, more carefully — for a transmission to happen, only one direction matters typically. Common approximation: factor of $(1-fp_1)$ once for transmitter (reduced shedding) and once for receiver (immunity). So:
\[
\beta_2 \to \beta_2(1-fp_1)^2
\]
Similarly recovery for vaccinated: $\gamma \to \gamma(1 + fp_2)$.

\textbf{When $f = 1$:}
\[
\beta_2^{\text{eff}} = \beta_2(1-p_1)^2, \quad \gamma^{\text{eff}} = \gamma(1+p_2)
\]
Variant spreads if:
\[
\beta_2(1-p_1)^2(N-S_2^*/2) > \gamma(1+p_2)
\]
$\iff (1-p_1)^2 R_{0,2} > 1 + p_2$\\
$\iff R_{0,2}^{\text{vacc}} = \frac{(1-p_1)^2 R_{0,2}}{1+p_2} > 1$.

Larger $p_1$ (better immunity) and larger $p_2$ (faster recovery) reduce variant spread. Vaccines help.
\end{solution}

\subsection{Problem 3.6: SEIR Model (Exposed Compartment)}

\begin{problem}
Add an \textbf{exposed} (latent) compartment $E$ to SIR. Infected individuals first pass through an exposed/incubating phase before becoming infectious:
\[
\frac{dS}{dt} = -\beta SI, \quad \frac{dE}{dt} = \beta SI - \sigma E, \quad \frac{dI}{dt} = \sigma E - \gamma I, \quad \frac{dR}{dt} = \gamma I
\]
where $\sigma$ is the rate of progressing from exposed to infectious (incubation rate).

\begin{enumerate}[label=(\alph*)]
\item Verify conservation and reduce.
\item Find the disease-free FP.
\item Compute $J^*$ at the disease-free FP.
\item Find eigenvalues and $R_0$.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Conservation:}

\textbf{Step 1.} Sum all four equations:
\[
\dot S + \dot E + \dot I + \dot R = (-\beta SI) + (\beta SI - \sigma E) + (\sigma E - \gamma I) + \gamma I
\]

\textbf{Step 2.} Cancel pairs: $-\beta SI + \beta SI = 0$, $-\sigma E + \sigma E = 0$, $-\gamma I + \gamma I = 0$:
\[
= 0 \checkmark
\]

$N = S + E + I + R$ is conserved. Eliminate $R$; work with $(S, E, I)$.

\medskip\textbf{(b) Disease-free FP:}

\textbf{Step 1.} $E^* = I^* = 0$, $S^* = N$, $R^* = 0$.

\medskip\textbf{(c) Jacobian at $(N, 0, 0)$:}

\textbf{Step 1.} Define $f = -\beta SI$, $g = \beta SI - \sigma E$, $h = \sigma E - \gamma I$. Compute partials.

\textbf{Step 2.} Partials of $f$:
\[
\partial f/\partial S = -\beta I, \quad \partial f/\partial E = 0, \quad \partial f/\partial I = -\beta S
\]

\textbf{Step 3.} Partials of $g$:
\[
\partial g/\partial S = \beta I, \quad \partial g/\partial E = -\sigma, \quad \partial g/\partial I = \beta S
\]

\textbf{Step 4.} Partials of $h$:
\[
\partial h/\partial S = 0, \quad \partial h/\partial E = \sigma, \quad \partial h/\partial I = -\gamma
\]

\textbf{Step 5.} Evaluate at $(S,E,I) = (N,0,0)$:
\[
J^* = \begin{pmatrix} 0 & 0 & -\beta N \\ 0 & -\sigma & \beta N \\ 0 & \sigma & -\gamma \end{pmatrix}
\]
\stepwhy{Setting $I=0$ zeros out all terms involving $\beta I$; substituting $S=N$}

\medskip\textbf{(d) Eigenvalues:}

\textbf{Step 1.} First column is all zeros: $\lambda_1 = 0$.

\textbf{Step 2.} Remaining 2$\times$2 sub-matrix (from $E, I$ rows):
\[
M = \begin{pmatrix} -\sigma & \beta N \\ \sigma & -\gamma \end{pmatrix}
\]

\textbf{Step 3.} Trace:
\[
\text{Tr}(M) = -\sigma - \gamma < 0
\]

\textbf{Step 4.} Determinant:
\[
\text{Det}(M) = (-\sigma)(-\gamma) - (\beta N)(\sigma) = \sigma\gamma - \beta N\sigma = \sigma(\gamma - \beta N)
\]

\textbf{Step 5.} For disease-free FP to be stable, need Det $> 0$:
\[
\sigma(\gamma - \beta N) > 0 \thus \gamma > \beta N \thus \frac{\beta N}{\gamma} < 1 \thus R_0 < 1
\]
\stepwhy{$\sigma > 0$ always, so sign of Det is determined by $\gamma - \beta N$}

\textbf{Step 6.} Define:
\[
\boxed{R_0 = \frac{\beta N}{\gamma}}
\]
Same formula as standard SIR! The exposed compartment adds a delay but does not change $R_0$.

\textbf{Step 7.} Eigenvalues via quadratic formula:
\[
\lambda_\pm = \frac{-(\sigma+\gamma) \pm \sqrt{(\sigma+\gamma)^2 - 4\sigma(\gamma - \beta N)}}{2}
\]

\textbf{Step 8.} Disease spreads iff Det $< 0$, i.e.\ $R_0 > 1$. When $R_0 > 1$: Det $< 0$, discriminant $> (\sigma+\gamma)^2$, so $\sqrt{} > \sigma+\gamma$, making $\lambda_+ > 0$. Disease invades.
\end{solution}

\subsection{Problem 3.7: SIR with Births and Deaths}

\begin{problem}
Add equal birth and death rates $\mu$ to the standard SIR:
\[
\frac{dS}{dt} = \mu N - \beta SI - \mu S
\]
\[
\frac{dI}{dt} = \beta SI - \gamma I - \mu I
\]
\[
\frac{dR}{dt} = \gamma I - \mu R
\]
where the $\mu N$ term is births (all born susceptible) and $-\mu X$ terms are deaths in each compartment.

\begin{enumerate}[label=(\alph*)]
\item Verify conservation (or lack thereof). Does $N$ change?
\item Find the disease-free FP.
\item Compute $J^*$, find eigenvalues, define $R_0$.
\item How does births/deaths change $R_0$ compared to standard SIR?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Conservation check:}

\textbf{Step 1.} Sum all three:
\[
\dot S + \dot I + \dot R = (\mu N - \beta SI - \mu S) + (\beta SI - \gamma I - \mu I) + (\gamma I - \mu R)
\]

\textbf{Step 2.} Collect:
\[
= \mu N - \mu(S + I + R) - \beta SI + \beta SI - \gamma I + \gamma I
\]

\textbf{Step 3.} Cancel pairs and use $S+I+R = N$:
\[
= \mu N - \mu N = 0 \checkmark
\]

\textbf{Step 4.} Total population $N$ is \textbf{still conserved} (births balance deaths). Eliminate $R$.

\medskip\textbf{(b) Disease-free FP:}

At $I^* = 0$: $\dot S = \mu N - \mu S^* = 0 \Rightarrow S^* = N$. So $(S^*, I^*) = (N, 0)$.

\medskip\textbf{(c) Jacobian:}

\textbf{Step 1.} Partials of $f = \mu N - \beta SI - \mu S$:
\[
\partial f/\partial S = -\beta I - \mu, \quad \partial f/\partial I = -\beta S
\]

\textbf{Step 2.} Partials of $g = \beta SI - \gamma I - \mu I = \beta SI - (\gamma+\mu)I$:
\[
\partial g/\partial S = \beta I, \quad \partial g/\partial I = \beta S - (\gamma+\mu)
\]

\textbf{Step 3.} Build and evaluate at $(N, 0)$:
\[
J^* = \begin{pmatrix} -\mu & -\beta N \\ 0 & \beta N - (\gamma+\mu) \end{pmatrix}
\]
\stepwhy{Setting $I=0$: $\partial f/\partial S = -\mu$; $\partial g/\partial S = 0$}

\textbf{Step 4.} Upper triangular $\Rightarrow$ eigenvalues are diagonal:
\[
\lambda_1 = -\mu < 0, \qquad \lambda_2 = \beta N - (\gamma+\mu)
\]

\textbf{Step 5.} Factor $\lambda_2$:
\[
\lambda_2 = (\gamma+\mu)\left(\frac{\beta N}{\gamma+\mu} - 1\right) = (\gamma+\mu)(R_0 - 1)
\]
where
\[
\boxed{R_0 = \frac{\beta N}{\gamma + \mu}}
\]

\medskip\textbf{(d) Effect of births/deaths:}

\textbf{Standard SIR:} $R_0 = \beta N/\gamma$.

\textbf{SIR with births/deaths:} $R_0 = \beta N/(\gamma + \mu)$.

\textbf{Step 1.} Since $\gamma + \mu > \gamma$, the new $R_0 < $ standard $R_0$.

\textbf{Step 2.} \textbf{Births/deaths make disease \emph{harder} to spread (smaller $R_0$)}. Why?

\textbf{Step 3.} The additional $\mu$ in the denominator reflects that infected individuals also die at rate $\mu$ before recovering. They are removed from the infectious pool both by recovery ($\gamma$) and by death ($\mu$). Effectively, the \textbf{average infectious period} is $1/(\gamma+\mu)$ rather than $1/\gamma$.

\textbf{Step 4.} Herd immunity threshold: $f^* = 1 - 1/R_0 = 1 - (\gamma+\mu)/(\beta N)$. Slightly lower than without deaths, meaning fewer people need to be vaccinated.
\end{solution}

\newpage
%==================================================================
\section{Michaelis-Menten Kinetics}
%==================================================================

\subsection{Problem 4.1: Standard QSSA Derivation}

\begin{problem}
\[
S + E \xrightleftharpoons[k_{-1}]{k_1} [SE] \xrightarrow{k_2} P + E
\]
Total enzyme $E_0 = E + [SE]$ conserved.
\begin{enumerate}[label=(\alph*)]
\item Write all four equations.
\item Apply QSSA. Solve for $[SE]$ in terms of $S$ and $E_0$.
\item Find $V_\text{max}$, $K_M$.
\item Show limits: $S \ll K_M$ and $S \gg K_M$.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Equations:}

The reactions:
\begin{itemize}
\item $S + E \to [SE]$ at rate $k_1 SE$ (consumes $S, E$; produces $[SE]$)
\item $[SE] \to S + E$ at rate $k_{-1}[SE]$ (reverse: consumes $[SE]$; produces $S, E$)
\item $[SE] \to P + E$ at rate $k_2[SE]$ (consumes $[SE]$; produces $P, E$)
\end{itemize}

\textbf{Step 1.} Build each ODE by listing all rates that produce or consume that species:
\[
\dot S = -k_1 SE + k_{-1}[SE]
\]
\stepwhy{$S$ consumed at rate $k_1 SE$; replenished at rate $k_{-1}[SE]$}
\[
\dot E = -k_1 SE + k_{-1}[SE] + k_2[SE] = -k_1 SE + (k_{-1}+k_2)[SE]
\]
\stepwhy{$E$ consumed forming complex; replenished by both back-reactions; combine like terms}
\[
\dot{[SE]} = k_1 SE - k_{-1}[SE] - k_2[SE] = k_1 SE - (k_{-1}+k_2)[SE]
\]
\[
\dot P = k_2[SE]
\]

\textbf{Step 2.} Verify conservation. Sum $\dot E + \dot{[SE]}$:
\[
[-k_1 SE + (k_{-1}+k_2)[SE]] + [k_1 SE - (k_{-1}+k_2)[SE]]
\]
\[
= -k_1 SE + k_1 SE + (k_{-1}+k_2)[SE] - (k_{-1}+k_2)[SE] = 0 \checkmark
\]
So $E + [SE] = E_0$ constant.

\medskip\textbf{(b) QSSA derivation:}

\textbf{Step 1.} Set $d[SE]/dt = 0$:
\[
k_1 SE - (k_{-1}+k_2)[SE] = 0
\]

\textbf{Step 2.} Move the second term to RHS:
\[
k_1 SE = (k_{-1}+k_2)[SE]
\]

\textbf{Step 3.} Use conservation $E = E_0 - [SE]$. Substitute:
\[
k_1 S(E_0 - [SE]) = (k_{-1}+k_2)[SE]
\]

\textbf{Step 4.} Distribute $k_1 S$ on LHS:
\[
k_1 S E_0 - k_1 S [SE] = (k_{-1}+k_2)[SE]
\]

\textbf{Step 5.} Move $-k_1 S[SE]$ to RHS:
\[
k_1 S E_0 = (k_{-1}+k_2)[SE] + k_1 S[SE]
\]

\textbf{Step 6.} Factor $[SE]$ on RHS:
\[
k_1 S E_0 = [SE]\cdot(k_{-1} + k_2 + k_1 S)
\]

\textbf{Step 7.} Divide both sides by $(k_{-1} + k_2 + k_1 S)$:
\[
[SE] = \frac{k_1 S E_0}{k_{-1} + k_2 + k_1 S}
\]

\textbf{Step 8.} Divide numerator and denominator by $k_1$:
\[
[SE] = \frac{S E_0}{(k_{-1}+k_2)/k_1 + S}
\]
\stepwhy{$k_1 S/k_1 = S$ in numerator; $k_1 S/k_1 = S$ in denominator; $(k_{-1}+k_2)/k_1$ remains}

\textbf{Step 9.} Define $K_M = (k_{-1}+k_2)/k_1$:
\[
\boxed{[SE] = \frac{E_0 S}{K_M + S}}
\]

\medskip\textbf{(c) Production rate:}

\textbf{Step 1.} From $\dot P = k_2[SE]$, substitute the QSSA result:
\[
\frac{dP}{dt} = k_2\cdot\frac{E_0 S}{K_M + S} = \frac{k_2 E_0 S}{K_M + S}
\]

\textbf{Step 2.} Define $V_{\max} = k_2 E_0$:
\[
\boxed{\frac{dP}{dt} = \frac{V_{\max} S}{K_M + S}}
\]

\medskip\textbf{(d) Limits:}

\textbf{Limit 1: $S \ll K_M$.} Denominator $K_M + S \approx K_M$:
\[
\frac{dP}{dt} \approx \frac{V_{\max} S}{K_M} = \frac{V_{\max}}{K_M}\cdot S
\]
Linear in $S$ (first-order kinetics). Slope is $V_{\max}/K_M$.

\textbf{Limit 2: $S \gg K_M$.} Denominator $K_M + S \approx S$:
\[
\frac{dP}{dt} \approx \frac{V_{\max} S}{S} = V_{\max}
\]
Constant (zero-order kinetics; saturation).

\textbf{At $S = K_M$:}
\[
\frac{dP}{dt} = \frac{V_{\max}\cdot K_M}{K_M + K_M} = \frac{V_{\max} K_M}{2K_M} = \frac{V_{\max}}{2}
\]
Half-maximum velocity (defining property).
\end{solution}

\subsection{Problem 4.2: Spike Protein Coupled Reaction (2023 Q4)}

\begin{problem}
\[
S + E \rightleftharpoons [SE] \to P + E
\]
\[
S + \tilde\sigma \rightleftharpoons [S\tilde\sigma] \to P_\text{new} + \tilde\sigma
\]
with rate constants $k_1, k_{-1}, k_2$ for the first reaction and $\tilde k_1, \tilde k_{-1}, \tilde k_2$ for the second. Spike protein $\tilde\sigma$ acts like a second enzyme.
\begin{enumerate}[label=(\alph*)]
\item Write all equations.
\item How many independent equations after conservation?
\item Apply QSSA. Find $\dot P$ and $\dot P_\text{new}$.
\item How would degradation modify this?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) All ODEs:}

The reactions involve $S$ binding both $E$ (rate $k_1$) and $\tilde\sigma$ (rate $\tilde k_1$), with both reactions producing products. Build each ODE by listing rates that consume or produce that species:

\[
\dot S = -k_1 SE + k_{-1}[SE] - \tilde k_1 S\tilde\sigma + \tilde k_{-1}[S\tilde\sigma]
\]
\stepwhy{$S$ consumed forming both complexes; replenished by both back-reactions}
\[
\dot E = -k_1 SE + (k_{-1}+k_2)[SE]
\]
\[
\dot{[SE]} = k_1 SE - (k_{-1}+k_2)[SE]
\]
\[
\dot{\tilde\sigma} = -\tilde k_1 S\tilde\sigma + (\tilde k_{-1}+\tilde k_2)[S\tilde\sigma]
\]
\[
\dot{[S\tilde\sigma]} = \tilde k_1 S\tilde\sigma - (\tilde k_{-1}+\tilde k_2)[S\tilde\sigma]
\]
\[
\dot P = k_2[SE], \qquad \dot P_{\text{new}} = \tilde k_2[S\tilde\sigma]
\]
That's 7 equations.

\medskip\textbf{(b) Conservation laws:}

\textbf{Step 1.} Verify first conservation: $\dot E + \dot{[SE]}$:
\[
[-k_1 SE + (k_{-1}+k_2)[SE]] + [k_1 SE - (k_{-1}+k_2)[SE]] = 0 \checkmark
\]
So $E + [SE] = E_0$.

\textbf{Step 2.} Verify second: $\dot{\tilde\sigma} + \dot{[S\tilde\sigma]}$:
\[
[-\tilde k_1 S\tilde\sigma + (\tilde k_{-1}+\tilde k_2)[S\tilde\sigma]] + [\tilde k_1 S\tilde\sigma - (\tilde k_{-1}+\tilde k_2)[S\tilde\sigma]] = 0 \checkmark
\]
So $\tilde\sigma + [S\tilde\sigma] = \tilde\sigma_0$.

\textbf{Step 3.} Two conservation laws $\Rightarrow$ eliminate $E$ and $\tilde\sigma$. Independent equations: $7 - 2 = 5$.

\medskip\textbf{(c) QSSA:}

\textbf{Step 1.} Apply QSSA to first complex: $\dot{[SE]} = 0$:
\[
k_1 SE = (k_{-1}+k_2)[SE]
\]

\textbf{Step 2.} Substitute $E = E_0 - [SE]$:
\[
k_1 S(E_0 - [SE]) = (k_{-1}+k_2)[SE]
\]

\textbf{Step 3.} Distribute and rearrange (same algebra as Problem 4.1):
\[
k_1 S E_0 = [SE](k_{-1} + k_2 + k_1 S)
\]

\textbf{Step 4.} Solve, divide by $k_1$:
\[
[SE] = \frac{S E_0}{K_M + S}, \quad K_M = \frac{k_{-1}+k_2}{k_1}
\]

\textbf{Step 5.} Same procedure for second complex $\dot{[S\tilde\sigma]} = 0$:
\[
\tilde k_1 S\tilde\sigma = (\tilde k_{-1}+\tilde k_2)[S\tilde\sigma]
\]

\textbf{Step 6.} Substitute $\tilde\sigma = \tilde\sigma_0 - [S\tilde\sigma]$:
\[
\tilde k_1 S(\tilde\sigma_0 - [S\tilde\sigma]) = (\tilde k_{-1}+\tilde k_2)[S\tilde\sigma]
\]

\textbf{Step 7.} Solve, divide by $\tilde k_1$:
\[
[S\tilde\sigma] = \frac{S\tilde\sigma_0}{\tilde K_M + S}, \quad \tilde K_M = \frac{\tilde k_{-1}+\tilde k_2}{\tilde k_1}
\]

\textbf{Step 8.} Production rates:
\[
\frac{dP}{dt} = k_2[SE] = \frac{k_2 E_0 S}{K_M + S}
\]
\[
\frac{dP_{\text{new}}}{dt} = \tilde k_2[S\tilde\sigma] = \frac{\tilde k_2\tilde\sigma_0 S}{\tilde K_M + S}
\]

\textbf{Important observation:} The two reactions \emph{compete} for the substrate $S$ (it's consumed by both pathways). But because $E_0$ and $\tilde\sigma_0$ are separately conserved, in QSSA each reaction's rate depends only on $S$ via its own MM expression. The competition shows up in how $S$ depletes over time, not in either rate directly.

\medskip\textbf{(d) With degradation:}

\textbf{Step 1.} Modify equations:
\[
\dot P = k_2[SE] - k_d P
\]

\textbf{Step 2.} Setting $\dot P = 0$ (steady state):
\[
k_2[SE] = k_d P^* \thus P^* = \frac{k_2[SE]}{k_d}
\]

\textbf{Step 3.} Substitute QSSA result for $[SE]$:
\[
P^* = \frac{k_2 E_0 S}{k_d(K_M + S)}
\]

\textbf{Step 4.} Now \textbf{both} $P$ and $P_{\text{new}}$ reach steady states (rather than growing forever).

\textbf{Step 5.} \textbf{Short-term behavior}: $P_{\text{new}}$ accumulates linearly at rate $\tilde k_2\tilde\sigma_0 S/(\tilde K_M + S)$.

\textbf{Step 6.} \textbf{Long-term behavior}: $P_{\text{new}}$ saturates at $P_{\text{new}}^* = \tilde k_2\tilde\sigma_0 S/[k_d(\tilde K_M + S)]$.

\textbf{Vaccine implications:} A potentially harmful product accumulates short-term but resolves long-term if degradation operates. If the byproduct triggers autoimmune feedback (additional rate boost), the system might not stabilize.
\end{solution}

\subsection{Problem 4.3: Hill Function and Cooperativity}

\begin{problem}
For a reaction with cooperativity $n$:
\[
\dot P = \frac{V_\text{max} S^n}{K^n + S^n}
\]
\begin{enumerate}[label=(\alph*)]
\item At what $S$ is the rate half-maximum?
\item Compute $\frac{d(\dot P)}{dS}$ at $S = K$. How does this depend on $n$?
\item Plot the difference between $n = 1$ and $n = 4$.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Half-max concentration:}

\textbf{Step 1.} Set $\dot P = V_{\max}/2$:
\[
\frac{V_{\max} S^n}{K^n + S^n} = \frac{V_{\max}}{2}
\]

\textbf{Step 2.} Cancel $V_{\max}$ from both sides:
\[
\frac{S^n}{K^n + S^n} = \frac{1}{2}
\]

\textbf{Step 3.} Cross-multiply:
\[
2S^n = K^n + S^n
\]

\textbf{Step 4.} Subtract $S^n$ from both sides:
\[
S^n = K^n
\]

\textbf{Step 5.} Take $n$-th root:
\[
S = K
\]

The half-maximum concentration is \textbf{$S = K$ regardless of $n$}.

\medskip\textbf{(b) Slope at half-max:}

\textbf{Step 1.} Compute $\frac{d(\dot P)}{dS}$ using the quotient rule. Let $u = V_{\max}S^n$ and $v = K^n + S^n$. Then $u' = V_{\max}\cdot nS^{n-1}$ and $v' = nS^{n-1}$:
\[
\frac{d(\dot P)}{dS} = \frac{u'v - uv'}{v^2}
\]

\textbf{Step 2.} Numerator:
\[
u'v - uv' = V_{\max}nS^{n-1}(K^n + S^n) - V_{\max}S^n\cdot nS^{n-1}
\]

\textbf{Step 3.} Factor $V_{\max}nS^{n-1}$ out:
\[
= V_{\max}nS^{n-1}[(K^n + S^n) - S^n] = V_{\max}nS^{n-1}\cdot K^n
\]
\stepwhy{$(K^n + S^n) - S^n = K^n$}

\textbf{Step 4.} So:
\[
\frac{d(\dot P)}{dS} = \frac{V_{\max}nK^n S^{n-1}}{(K^n + S^n)^2}
\]

\textbf{Step 5.} Evaluate at $S = K$. First, $S^{n-1} = K^{n-1}$:
\[
\frac{d(\dot P)}{dS}\bigg|_{S=K} = \frac{V_{\max}nK^n\cdot K^{n-1}}{(K^n + K^n)^2}
\]

\textbf{Step 6.} Simplify numerator: $K^n\cdot K^{n-1} = K^{2n-1}$:
\[
= \frac{V_{\max}n K^{2n-1}}{(2K^n)^2}
\]

\textbf{Step 7.} Simplify denominator: $(2K^n)^2 = 4K^{2n}$:
\[
= \frac{V_{\max}n K^{2n-1}}{4K^{2n}}
\]

\textbf{Step 8.} Cancel $K^{2n-1}/K^{2n} = 1/K$:
\[
= \frac{V_{\max}n}{4K}
\]
\[
\boxed{\frac{d(\dot P)}{dS}\bigg|_{S=K} = \frac{nV_{\max}}{4K}}
\]

\textbf{Slope at half-max scales linearly with $n$.} Higher $n$ means steeper switch.

\medskip\textbf{(c) Comparison of $n=1$ vs $n=4$:}

\begin{itemize}
\item \textbf{$n=1$ (standard MM):} smooth hyperbolic increase. No inflection point.
\item \textbf{$n=4$ (high cooperativity):} switch-like sigmoid. Very flat at low $S$, sharp rise near $S = K$, then flat at high $S$. Inflection point at $S = K$.
\end{itemize}

Biologically, cooperative binding (high $n$) produces \textbf{``digital'' threshold response}: gene either off or on, channel either closed or open, near a critical concentration. This is the basis of switching in many cell-signaling pathways.
\end{solution}

\newpage
%==================================================================
\section{Hodgkin-Huxley and Action Potentials}
%==================================================================

\subsection{Problem 5.1: Conceptual Questions on HH}

\begin{problem}
\begin{enumerate}[label=(\alph*)]
\item Write the HH equation for membrane voltage.
\item Explain each term physically.
\item Why $m^3 h$ for sodium and $n^4$ for potassium?
\item Sketch dynamics of $V$, $m$, $h$, $n$ during an action potential.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a)}
\[
C_m \frac{dV}{dt} = -g_\text{Na} m^3 h(V - V_\text{Na}) - g_K n^4(V - V_K) - g_L(V - V_L) + I_\text{ext}
\]

\textbf{(b) Terms:}
\begin{itemize}
\item $C_m \dot V$: capacitive current (charging/discharging the membrane).
\item $g_\text{Na} m^3 h$: maximum Na conductance times fraction of channels open ($m^3 h$).
\item $(V - V_\text{Na})$: driving force; current = conductance $\times$ driving force.
\item $g_L (V - V_L)$: passive ``leak'' through always-open channels.
\item $I_\text{ext}$: applied stimulus.
\end{itemize}

\textbf{(c) Powers:} 
\begin{itemize}
\item Sodium has 3 activation gates (each opens with prob $m$), and 1 inactivation gate (closes with prob $1-h$, so open with $h$). Net open: $m^3 h$.
\item Potassium has 4 activation gates: $n^4$.
\item These powers come from the structural biology of the channels (subunits).
\end{itemize}

\textbf{(d) Action potential phases:}
\begin{enumerate}
\item \textbf{Resting} ($V \approx -70$ mV): $m \approx 0$, $h \approx 1$, $n \approx 0$.
\item \textbf{Depolarization}: stimulus $\to V \uparrow \to m \uparrow$ rapidly $\to$ Na current floods in $\to V \uparrow$ further (positive feedback).
\item \textbf{Peak} ($V \approx +30$ mV): $m$ saturates at 1.
\item \textbf{Repolarization}: $h \downarrow$ slowly (Na inactivates) and $n \uparrow$ slowly (K opens). Na current shuts off; K current flows out.
\item \textbf{Hyperpolarization (afterpolarization)}: K current overshoots; $V$ briefly below resting.
\item \textbf{Recovery}: $h$ slowly returns to 1; $n$ to 0; refractory period.
\end{enumerate}
\end{solution}

\subsection{Problem 5.2: Gating Variable Dynamics}

\begin{problem}
A gating variable $x$ obeys:
\[
\frac{dx}{dt} = \alpha_x(V)(1-x) - \beta_x(V) x
\]
\begin{enumerate}[label=(\alph*)]
\item Find steady-state $x_\infty(V)$ and time constant $\tau_x(V)$.
\item Rewrite the ODE in standard relaxation form.
\item Solve for $x(t)$ if $V$ is held constant after $t=0$.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Steady-state and time constant:}

\textbf{Step 1.} At steady state, $\dot x = 0$:
\[
\alpha_x(1-x_\infty) - \beta_x x_\infty = 0
\]

\textbf{Step 2.} Distribute $\alpha_x$:
\[
\alpha_x - \alpha_x x_\infty - \beta_x x_\infty = 0
\]

\textbf{Step 3.} Group $x_\infty$ terms:
\[
\alpha_x = \alpha_x x_\infty + \beta_x x_\infty = x_\infty(\alpha_x + \beta_x)
\]

\textbf{Step 4.} Solve for $x_\infty$:
\[
\boxed{x_\infty(V) = \frac{\alpha_x(V)}{\alpha_x(V) + \beta_x(V)}}
\]

\textbf{Step 5.} For $\tau_x$, expand the original ODE:
\[
\dot x = \alpha_x(1-x) - \beta_x x = \alpha_x - \alpha_x x - \beta_x x = \alpha_x - (\alpha_x + \beta_x) x
\]
\stepwhy{distribute, then factor $-x$ from last two terms}

\textbf{Step 6.} Compare with the standard relaxation form $\dot x = (x_\infty - x)/\tau_x$. Reading off:
\[
\boxed{\tau_x(V) = \frac{1}{\alpha_x(V) + \beta_x(V)}}
\]

\medskip\textbf{(b) Standard relaxation form:}

\textbf{Step 1.} From step 5: $\dot x = \alpha_x - (\alpha_x + \beta_x)x$. Factor $(\alpha_x + \beta_x)$:
\[
\dot x = (\alpha_x + \beta_x)\left[\frac{\alpha_x}{\alpha_x + \beta_x} - x\right]
\]

\textbf{Step 2.} Recognize $\alpha_x/(\alpha_x + \beta_x) = x_\infty$ and $(\alpha_x + \beta_x) = 1/\tau_x$:
\[
\boxed{\frac{dx}{dt} = \frac{x_\infty - x}{\tau_x}}
\]

\medskip\textbf{(c) Solution for constant $V$:}

\textbf{Step 1.} If $V$ is constant, $\alpha_x$, $\beta_x$, $x_\infty$, $\tau_x$ are all constants. Define $y = x - x_\infty$, so $\dot y = \dot x$:
\[
\dot y = \frac{x_\infty - (y + x_\infty)}{\tau_x} = \frac{-y}{\tau_x}
\]

\textbf{Step 2.} This is exponential decay:
\[
\frac{dy}{y} = -\frac{dt}{\tau_x} \thus \int\frac{dy}{y} = -\int\frac{dt}{\tau_x} \thus \ln|y| = -\frac{t}{\tau_x} + C
\]

\textbf{Step 3.} Exponentiate:
\[
y(t) = y(0)\,e^{-t/\tau_x}
\]
\stepwhy{$y(0) = e^C$}

\textbf{Step 4.} Substitute back $y = x - x_\infty$:
\[
x(t) - x_\infty = (x(0) - x_\infty)\,e^{-t/\tau_x}
\]

\textbf{Step 5.}
\[
\boxed{x(t) = x_\infty + (x(0) - x_\infty)\,e^{-t/\tau_x}}
\]

Exponential approach to $x_\infty$ with characteristic time $\tau_x$. As $t \to \infty$, $e^{-t/\tau_x} \to 0$, so $x \to x_\infty$. \checkmark
\end{solution}

\newpage
%==================================================================
\section{Stability \& Eigenvalue Analysis (General)}
%==================================================================

\subsection{Problem 6.1: 2$\times$2 Matrix Eigenvalue Practice}

\begin{problem}
For $M = \begin{pmatrix} 2 & -1 \\ 4 & -3 \end{pmatrix}$, compute Tr, Det, and eigenvalues. Classify the fixed point.
\end{problem}

\begin{solution}
\textbf{Step 1.} Identify entries: $a = 2$, $b = -1$, $c = 4$, $d = -3$.

\textbf{Step 2.} Trace = sum of diagonal:
\[
\text{Tr}(M) = a + d = 2 + (-3) = 2 - 3 = -1
\]

\textbf{Step 3.} Determinant = $ad - bc$:
\[
\text{Det}(M) = (2)(-3) - (-1)(4) = -6 - (-4) = -6 + 4 = -2
\]
\stepwhy{$-(-4) = +4$}

\textbf{Step 4.} Compute discriminant:
\[
\Delta = \text{Tr}^2 - 4\text{Det} = (-1)^2 - 4(-2) = 1 - (-8) = 1 + 8 = 9
\]

\textbf{Step 5.} Apply quadratic formula:
\[
\lambda_\pm = \frac{\text{Tr} \pm \sqrt\Delta}{2} = \frac{-1 \pm \sqrt 9}{2} = \frac{-1 \pm 3}{2}
\]

\textbf{Step 6.} Compute both:
\[
\lambda_+ = \frac{-1 + 3}{2} = \frac{2}{2} = 1
\]
\[
\lambda_- = \frac{-1 - 3}{2} = \frac{-4}{2} = -2
\]

\textbf{Step 7.} Verify with characteristic polynomial $\lambda^2 - \text{Tr}\lambda + \text{Det} = 0$:
\[
\lambda^2 + \lambda - 2 = 0
\]
Factor: $(\lambda + 2)(\lambda - 1) = 0$, giving $\lambda = -2$ or $\lambda = 1$. \checkmark

\textbf{Step 8.} \textbf{Classification:} One positive ($\lambda_+ = 1$), one negative ($\lambda_- = -2$). This is a \textbf{saddle point} (always unstable).

Det $< 0$ on its own would have told us it's a saddle (the row in the stability table for ``any Tr, Det $< 0$'').
\end{solution}

\subsection{Problem 6.2: Asymptotic Analysis of a Perturbation (2024 Q4)}

\begin{problem}
A perturbation $\delta x(t)$ obeys:
\[
\delta x(t) = \frac{1}{1 - \frac{1}{\lambda\tau}e^{\lambda t}/\lambda} \cdot \frac{1}{\tau}
\]
\begin{enumerate}[label=(\alph*)]
\item Behavior at early times $t \ll \tau$.
\item Behavior at long times $t \gg \tau$.
\item Sketch in phase space.
\item If this is only a leading-order Taylor expansion, what's the conclusion?
\item For a different system: $\delta x(t) = (1 + e^{-\lambda t})e^{-\lambda t}$. Long-time behavior?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Early times:} $e^{\lambda t}$ near 1, so denominator $\approx 1 - 1/(\lambda^2 \tau) \approx 1$ if $\lambda^2 \tau \gg 1$. Leading behavior $\delta x \approx 1/\tau \cdot e^{\lambda t}/\text{stuff}$. More carefully: at small $t$, expand exponential, $\delta x \approx \frac{1}{\tau} \cdot e^{\lambda t}$ (exponential growth).

\textbf{(b) Long times:} $e^{\lambda t} \to \infty$ (since $\lambda > 0$), denominator $\to -\infty$ in magnitude, so $\delta x \to 0$. Perturbation \textbf{decays back to FP eventually}.

\textbf{(c) Phase space:} Trajectory first moves away from FP, then turns around and comes back. Looks like a loop or excursion.

\textbf{Conclusion:} Since the perturbation eventually decays to zero, the system is \textbf{stable} (this is the exact solution).

\textbf{(d) If only leading-order Taylor:} Cannot conclude. Higher-order terms might dominate when the perturbation grows large; the linear theory only valid near FP. The trajectory might escape before higher orders bring it back, leading to true instability.

\textbf{(e) Other system:} $\delta x(t) = (1 + e^{-\lambda t})e^{-\lambda t} = e^{-\lambda t} + e^{-2\lambda t} \to 0$ as $t \to \infty$. Stable, with no oscillations (real exponentials).

If $\lambda$ is complex, say $\lambda = a + i\omega$, then $e^{-\lambda t}$ oscillates with envelope $e^{-at}$. If $a > 0$, decays $\Rightarrow$ \textbf{spiral into FP}. Stable spiral (limit-cycle-like decay).
\end{solution}

\subsection{Problem 6.3: 3$\times$3 Stability via Routh-Hurwitz}

\begin{problem}
For 3$\times$3 matrix $J^*$, when is the FP stable?
\end{problem}

\begin{solution}
Characteristic polynomial: $\lambda^3 + a_2\lambda^2 + a_1 \lambda + a_0 = 0$ where:
\begin{itemize}
\item $a_2 = -\text{Tr}(J^*)$
\item $a_1 = $ sum of 2$\times$2 principal minors
\item $a_0 = -\text{Det}(J^*)$
\end{itemize}
\textbf{Routh-Hurwitz conditions for all roots to have negative real parts:}
\begin{enumerate}
\item $a_2 > 0$ (i.e.\ $\text{Tr}(J^*) < 0$).
\item $a_0 > 0$ (i.e.\ $\text{Det}(J^*) < 0$).
\item $a_2 a_1 > a_0$.
\end{enumerate}

\textbf{Number of invariants for $n \times n$:} For $n \times n$, you need $n$ independent equations (Tr, Det, and $n-2$ intermediate invariants for $n > 2$). Always $n-2$ more than 2$\times$2 case.
\end{solution}

\subsection{Problem 6.4: Trace and Determinant Biological Meaning (2019 Q2)}

\begin{problem}
For each matrix below, describe the biological meaning of Tr, Det, and predict stability.
\[
A_1 = \begin{pmatrix} -r/K & -\bar\beta \\ \varepsilon\bar\beta & 0 \end{pmatrix}, \quad
A_2 = \begin{pmatrix} 0 & -\beta \\ \beta & 0 \end{pmatrix}
\]
\end{problem}

\begin{solution}
\textbf{$A_1$ (logistic-prey LV):}
\begin{itemize}
\item $\text{Tr}(A_1) = -r/K$ (sum of self-interactions). Only prey self-regulates (logistic). Negative $\Rightarrow$ stabilizing.
\item $\text{Det}(A_1) = \varepsilon\bar\beta^2$ (product of cross-species interactions). Positive $\Rightarrow$ stabilizing form (no saddle).
\item Eigenvalues: $\lambda_\pm = -\frac{r}{2K}\left(1 \pm \sqrt{1 - \frac{4\varepsilon\bar\beta^2 K^2}{r^2}}\right)$. Real part negative $\Rightarrow$ \textbf{stable}.
\end{itemize}

\textbf{$A_2$ (basic SIR):}
\begin{itemize}
\item $\text{Tr}(A_2) = 0$: no self-interactions among compartments.
\item $\text{Det}(A_2) = \beta^2 > 0$: positive cross-interactions $\Rightarrow$ oscillatory.
\item $\lambda_\pm = \pm i\beta$: pure imaginary, neutral oscillations.
\end{itemize}
\end{solution}

\newpage
%==================================================================
\section{Conservation Equations \& Constraint Surfaces}
%==================================================================

\subsection{Problem 7.1: Identifying Conservation}

\begin{problem}
For each system, determine whether $N = \sum X_i$ is conserved:
\begin{enumerate}[label=(\alph*)]
\item SIR with deaths: $\dot S = -\beta SI$, $\dot I = \beta SI - \gamma I - \mu I$, $\dot R = \gamma I$.
\item SIRS: $\dot S = -\beta SI + \nu R$, $\dot I = \beta SI - \gamma I$, $\dot R = \gamma I - \nu R$.
\item Logistic predator-prey.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a)} 

\textbf{Step 1.} Sum all three: $\dot S + \dot I + \dot R = (-\beta SI) + (\beta SI - \gamma I - \mu I) + (\gamma I)$.

\textbf{Step 2.} Cancel: $-\beta SI + \beta SI = 0$, $\gamma I - \gamma I = 0$. Remaining: $-\mu I$.

\textbf{Step 3.} So $\frac{d(S+I+R)}{dt} = -\mu I \leq 0$. \textbf{Not conserved}; total population decreases due to disease deaths. Cannot eliminate any equation.

\textbf{(b)}

\textbf{Step 1.} Sum: $(-\beta SI + \nu R) + (\beta SI - \gamma I) + (\gamma I - \nu R)$.

\textbf{Step 2.} Cancel: $-\beta SI + \beta SI = 0$, $-\gamma I + \gamma I = 0$, $\nu R - \nu R = 0$. Result: 0.

\textbf{Step 3.} \textbf{Conserved}. $N = S+I+R = $ const. Eliminate one equation (e.g., $R = N - S - I$).

\textbf{(c)}

\textbf{Step 1.} Sum: $(rN(1-N/K) - \bar\beta NP) + (\varepsilon\bar\beta NP - ZP)$.

\textbf{Step 2.} Collect: $rN - rN^2/K - \bar\beta NP + \varepsilon\bar\beta NP - ZP = rN - rN^2/K + (\varepsilon-1)\bar\beta NP - ZP$.

\textbf{Step 3.} This is not zero (has leftover terms). \textbf{Not conserved}. Predator-prey has no biomass conservation because $\varepsilon < 1$ means prey biomass is lost as heat, and $-ZP$ represents death.
\end{solution}

\subsection{Problem 7.2: Reducing a 5-Variable System}

\begin{problem}
For the two-variant disease in Problem 3.4, show that conservation reduces 5 variables to 4 and explain why we can't reduce further despite there being multiple constraints.
\end{problem}

\begin{solution}
\textbf{Step 1.} The system has $S, I, S_2, I_2, R$ — five variables.

\textbf{Step 2.} Sum all five equations. Every term appears with both a positive and negative sign across the five equations (e.g., $\beta SI$ appears $-$ in $\dot S$ and $+$ in $\dot I$). Net sum: 0.

\textbf{Step 3.} Therefore $S + I + S_2 + I_2 + R = N$ is conserved. This lets us eliminate one variable: $R = N - S - I - S_2 - I_2$. Now 4 independent equations.

\textbf{Step 4.} Can we reduce further? Only if there is a \textbf{second independent conservation law}. Check: is there any other linear combination $c_1 S + c_2 I + c_3 S_2 + c_4 I_2 = $ const?

\textbf{Step 5.} For $c_1 \dot S + c_2 \dot I + c_3 \dot{S_2} + c_4 \dot{I_2} = 0$ to hold identically, we need the coefficients of each product ($SI$, $SI_2$, $S_2 I_2$) to cancel. This only works for $c_1 = c_2 = c_3 = c_4 = 1$ (total conservation). No other independent combination gives zero.

\textbf{Conclusion:} The system has exactly one conservation law and reduces from 5 to 4 independent equations. Cannot reduce further.
\end{solution}

\subsection{Problem 7.3: Enzyme Conservation with Inhibitor (2018 Q4 style)}

\begin{problem}
Consider competitive inhibition:
\[
S + E \rightleftharpoons [SE] \to P + E, \qquad I + E \rightleftharpoons [EI]
\]
\begin{enumerate}[label=(\alph*)]
\item Identify all conservation laws.
\item Apply QSSA to both complexes.
\item Derive the modified production rate. How does inhibitor affect $V_{\max}$ and $K_M$?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Conservation laws:}

\textbf{Step 1.} Sum $\dot E + \dot{[SE]} + \dot{[EI]}$. Using the full equations:
\[
\dot E = -k_1 SE + (k_{-1}+k_2)[SE] - k_3 IE + k_{-3}[EI]
\]
\[
\dot{[SE]} = k_1 SE - (k_{-1}+k_2)[SE]
\]
\[
\dot{[EI]} = k_3 IE - k_{-3}[EI]
\]
Sum: all terms cancel. So $E + [SE] + [EI] = E_0$ conserved.

\textbf{Step 2.} No conservation on $S$ or $I$ (they are consumed/released without matching terms).

\textbf{Step 3.} One conservation law: $E_0 = E + [SE] + [EI]$.

\medskip\textbf{(b) QSSA on both complexes:}

\textbf{Step 1.} Set $\dot{[SE]} = 0$:
\[
k_1 SE = (k_{-1}+k_2)[SE]
\]
\stepwhy{same as standard MM derivation}

\textbf{Step 2.} Set $\dot{[EI]} = 0$:
\[
k_3 IE = k_{-3}[EI] \thus [EI] = \frac{k_3 IE}{k_{-3}} = \frac{IE}{K_I}
\]
where $K_I = k_{-3}/k_3$ is the inhibitor dissociation constant.

\textbf{Step 3.} From Step 1: $[SE] = k_1 SE/(k_{-1}+k_2) = SE/K_M$ where $K_M = (k_{-1}+k_2)/k_1$.

\textbf{Step 4.} Substitute both into conservation $E + [SE] + [EI] = E_0$:
\[
E + \frac{SE}{K_M} + \frac{IE}{K_I} = E_0
\]

\textbf{Step 5.} Factor $E$:
\[
E\left(1 + \frac{S}{K_M} + \frac{I}{K_I}\right) = E_0 \thus E = \frac{E_0}{1 + S/K_M + I/K_I}
\]

\textbf{Step 6.} Find $[SE]$:
\[
[SE] = \frac{SE}{K_M} = \frac{S}{K_M}\cdot\frac{E_0}{1 + S/K_M + I/K_I}
\]

\textbf{Step 7.} Multiply top and bottom by $K_M$:
\[
[SE] = \frac{SE_0}{K_M + S + SK_M/K_I\cdot\frac{1}{S}\cdot S} = \frac{SE_0}{K_M(1+I/K_I) + S}
\]
Let me redo this cleanly:
\[
[SE] = \frac{S E_0 / K_M}{1 + S/K_M + I/K_I}
\]
Multiply numerator and denominator by $K_M$:
\[
[SE] = \frac{S E_0}{K_M + S + K_M I/K_I} = \frac{SE_0}{K_M(1 + I/K_I) + S}
\]

\medskip\textbf{(c) Modified production rate:}

\textbf{Step 1.}
\[
\dot P = k_2[SE] = \frac{k_2 E_0 S}{K_M(1+I/K_I) + S}
\]

\textbf{Step 2.} Compare to standard MM $\dot P = V_{\max}S/(K_M + S)$:
\begin{itemize}
\item $V_{\max}$ is \textbf{unchanged}: $V_{\max} = k_2 E_0$.
\item $K_M$ is \textbf{increased}: effective $K_M^{\text{eff}} = K_M(1 + I/K_I)$.
\end{itemize}

\textbf{Step 3.} Biological meaning: the inhibitor competes with substrate $S$ for the enzyme. At saturation ($S \gg K_M$), the reaction still reaches $V_{\max}$ (enough substrate displaces inhibitor). But at low $S$, the inhibitor effectively reduces affinity (larger apparent $K_M$ means enzyme needs more $S$ to reach half-max).
\end{solution}

\newpage
%==================================================================
\section{Mixed / Compound Problems}
%==================================================================

\subsection{Problem 8.1: COVID-19 Real-World Application (2020 Q1-3)}

\begin{problem}
A pathogen doubles cases every 5 days (factor of $e \approx 2.72$).
\begin{enumerate}[label=(\alph*)]
\item Find continuous growth rate $r$.
\item If 0.5\% of cases are fatal with 3-week delay, find deaths $D(t)$ on day $t$.
\item Tests are processed with a 3-day delay; reported cases $R(t)$?
\item Evaluate at $t = 100$ days starting from $N(0) = 1$.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a)} $N(t) = e^{rt}$ with $N(5) = e$ requires $5r = 1 \Rightarrow r = 0.2$ /day.

\textbf{(b)} Cases at time $t - 21$ days (3 weeks earlier) become deaths today, with 0.5\% probability:
\[
D(t) = 0.005 \cdot N(t-21) = 0.005 \cdot e^{0.2(t-21)} = 0.005\,e^{-4.2}\,e^{0.2t} \approx 7.5 \times 10^{-5} e^{0.2t}
\]

\textbf{(c)} Reported cases lag by 3 days:
\[
R(t) = N(t-3) = e^{0.2(t-3)} = e^{-0.6} e^{0.2t} \approx 0.549 e^{0.2t}
\]
New reported cases per day: $\Delta R(t) = R(t) - R(t-1) \approx R(t)(1 - e^{-0.2}) \approx 0.181 R(t)$.

\textbf{(d) At $t = 100$:}
\begin{itemize}
\item $N(100) = e^{20} \approx 4.85 \times 10^8$ (unrealistic; suggests population finite, model breakdown).
\item $D(100) = 7.5 \times 10^{-5} \cdot e^{20} \approx 36{,}000$
\item $R(100) \approx 0.549 \cdot e^{20} \approx 2.66 \times 10^8$
\end{itemize}
The numbers are unrealistic because exponential growth ignores the finite population — this is why we need SIR-type models with saturation.
\end{solution}

\subsection{Problem 8.2: Connecting SI Model to Exponential Growth (2020 Q2)}

\begin{problem}
For the SI model, $\dot I = \beta S I - \gamma I$ with $S + I = T$ (total constant). Show how, when $I \ll T$, this reduces to exponential growth. What is the effective $r$?
\end{problem}

\begin{solution}
\textbf{Step 1.} Use conservation: $S + I = T$ implies $S = T - I$.

\textbf{Step 2.} Substitute into $\dot I$:
\[
\dot I = \beta(T - I)I - \gamma I
\]

\textbf{Step 3.} Distribute $\beta(T-I)$:
\[
\dot I = \beta TI - \beta I^2 - \gamma I
\]

\textbf{Step 4.} Approximation: when $I \ll T$, the term $\beta I^2$ is negligible compared to $\beta TI$ (since $I^2 \ll TI$):
\[
\dot I \approx \beta TI - \gamma I
\]

\textbf{Step 5.} Factor $I$:
\[
\dot I = I(\beta T - \gamma)
\]

\textbf{Step 6.} Recognize this as exponential growth $\dot I = rI$ with:
\[
\boxed{r = \beta T - \gamma}
\]

\textbf{Step 7.} Express in terms of $R_0$. Factor $\gamma$:
\[
r = \gamma\left(\frac{\beta T}{\gamma} - 1\right) = \gamma(R_0 - 1)
\]

\textbf{Step 8.} Conclusions:
\begin{itemize}
\item $R_0 > 1 \Rightarrow r > 0 \Rightarrow$ exponential growth (early epidemic).
\item $R_0 < 1 \Rightarrow r < 0 \Rightarrow$ exponential decay (disease dies out).
\item $R_0 = 1 \Rightarrow r = 0 \Rightarrow$ marginal case.
\end{itemize}

\medskip\textbf{Without approximation:}

\textbf{Step 1.} Keep full equation $\dot I = \beta TI - \beta I^2 - \gamma I$:
\[
\dot I = (\beta T - \gamma)I - \beta I^2
\]

\textbf{Step 2.} Factor:
\[
\dot I = I[(\beta T - \gamma) - \beta I]
\]

\textbf{Step 3.} Compare to logistic form $\dot N = rN(1 - N/K)$ where $r = \beta T - \gamma$ and $rN/K = \beta I^2$, giving:
\[
\frac{r}{K} = \beta \thus K = \frac{r}{\beta} = \frac{\beta T - \gamma}{\beta}
\]

\textbf{Step 4.} Simplify:
\[
K = T - \frac{\gamma}{\beta} = T\left(1 - \frac{1}{R_0}\right)
\]
\stepwhy{factor $T$, recognize $\gamma/(\beta T) = 1/R_0$}

\textbf{Step 5.} \textbf{Conclusion:} The SI model is a logistic equation with carrying capacity $K = T(1-1/R_0)$. As the epidemic progresses, $I$ saturates at this fraction of the population.
\end{solution}

\subsection{Problem 8.3: Comparing Two Cities (2020 Q2b)}

\begin{problem}
City A: population $T_A$, infection rate $\beta_A$, $R_{0,A} = 1.1$. City B: $10\times$ population, $5\times$ infection rate. Compare long-time infected fraction.
\end{problem}

\begin{solution}
\textbf{Step 1.} Use $R_0 = \beta N/\gamma$ (where $N = T$ here) for both cities.

\textbf{Step 2.} City A: $R_{0,A} = \beta_A T_A/\gamma = 1.1$.

\textbf{Step 3.} City B parameters:
\[
T_B = 10\,T_A, \quad \beta_B = 5\,\beta_A
\]

\textbf{Step 4.} Compute $R_{0,B}$:
\[
R_{0,B} = \frac{\beta_B T_B}{\gamma} = \frac{(5\beta_A)(10 T_A)}{\gamma} = 50\cdot\frac{\beta_A T_A}{\gamma}
\]

\textbf{Step 5.} Substitute $R_{0,A} = 1.1$:
\[
R_{0,B} = 50\cdot R_{0,A} = 50\cdot 1.1 = 55
\]

\textbf{Massive amplification.}

\medskip\textbf{Long-time infected fraction:}

\textbf{Step 6.} For SIR, the final size equation is:
\[
1 - x = e^{-R_0 x}
\]
where $x = $ fraction eventually infected. Solve transcendentally.

\textbf{Step 7.} For City A ($R_0 = 1.1$): try $x = 0.18$:
\[
1 - 0.18 = 0.82, \quad e^{-1.1\cdot 0.18} = e^{-0.198} \approx 0.820 \checkmark
\]
So $x_A \approx 0.18$ (18\% infected).

\textbf{Step 8.} For City B ($R_0 = 55$): for any reasonable $x$ near 1:
\[
e^{-55\cdot x} \to 0 \text{ rapidly}
\]
so $1 - x_B \to 0$, meaning $x_B \to 1$ (essentially everyone infected).

\textbf{Step 9.} Comparison:
\begin{itemize}
\item Per capita: City A has 18\% infected; City B has $\sim 100\%$. Roughly $5\times$ more cases per capita.
\item Total cases: City A has $0.18 T_A$; City B has $\sim 10 T_A$ (since $T_B = 10T_A$). About $55\times$ more total cases.
\item \textbf{Hospital overflow risk}: City B will face catastrophic case load.
\end{itemize}

\textbf{Step 10.} Lesson: $R_0$ is super-linear in both transmission rate and population density. Doubling either creates much greater than 2× impact when it pushes $R_0$ above 1.
\end{solution}

\newpage
%==================================================================
\section{Model Formulation From Words (Pattern Recognition)}
%==================================================================

\subsection{Problem 9.1: Translating Biology to Equations}

\begin{problem}
For each scenario, write down the differential equations.
\begin{enumerate}[label=(\alph*)]
\item Two species in symbiosis (mutualism): both grow logistically, plus each benefits proportionally from the other.
\item A virus that mutates from strain 1 to strain 2 at rate $\mu$, with separate SIR dynamics for each.
\item Enzyme inhibition: an inhibitor $I$ binds the enzyme $E$ to form $[EI]$, reducing the active enzyme pool.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Symbiosis:}
\[
\dot N_1 = r_1 N_1 (1 - N_1/K_1) + \alpha_{12} N_1 N_2
\]
\[
\dot N_2 = r_2 N_2 (1 - N_2/K_2) + \alpha_{21} N_1 N_2
\]

\textbf{(b) Mutating virus (with conversion of strain 1 to 2 in infected hosts):}
\[
\dot S = -\beta_1 S I_1 - \beta_2 S I_2
\]
\[
\dot I_1 = \beta_1 S I_1 - \gamma I_1 - \mu I_1
\]
\[
\dot I_2 = \beta_2 S I_2 - \gamma I_2 + \mu I_1
\]
\[
\dot R = \gamma(I_1 + I_2)
\]

\textbf{(c) Enzyme inhibition (competitive):}
\[
S + E \rightleftharpoons [SE] \to P + E
\]
\[
I + E \rightleftharpoons [EI]
\]
Equations:
\[
\dot S = -k_1 SE + k_{-1}[SE]
\]
\[
\dot I = -k_3 IE + k_{-3}[EI]
\]
\[
\dot E = -k_1 SE + (k_{-1}+k_2)[SE] - k_3 IE + k_{-3}[EI]
\]
\[
\dot{[SE]} = k_1 SE - (k_{-1}+k_2)[SE]
\]
\[
\dot{[EI]} = k_3 IE - k_{-3}[EI]
\]
Conservation: $E_0 = E + [SE] + [EI]$. After QSSA on both complexes: $V_\text{max}^\text{eff} = V_\text{max}/(1 + I/K_I)$ where $K_I = k_{-3}/k_3$. Inhibitor reduces effective $V_\text{max}$.
\end{solution}

\subsection{Problem 9.2: Carrying Capacity Modification (2018 Q3)}

\begin{problem}
Standard LV without predator carrying capacity. What assumption underlies this?
What if the assumption fails? Modify equations.
\end{problem}

\begin{solution}
\textbf{Assumption:} Predator population is solely limited by prey availability (food). When prey are abundant, predators grow without saturation.

\textbf{When this fails:} Territoriality, intraspecific competition, mate-finding limits, or alternative resources.

\textbf{Modification:} Replace $\varepsilon\bar\beta N$ (per-capita predator growth) with $\varepsilon\bar\beta N(1 - P/K_P)$:
\[
\frac{dP}{dt} = \varepsilon\bar\beta NP(1 - P/K_P) - ZP
\]
Now $P$ has its own carrying capacity. Adds negative diagonal term to $A$:
\[
A = \begin{pmatrix} 0 & -\bar\beta \\ \varepsilon\bar\beta & -\varepsilon\bar\beta N^*/K_P \end{pmatrix}
\]
At fixed point, this becomes stabilizing.
\end{solution}

\subsection{Problem 9.3: Chronic Carriers (2018 Q4)}

\begin{problem}
SIR model with chronic carriers $C$:
\[
\dot S = -\beta SI - \alpha SC + \gamma I
\]
\[
\dot I = \beta SI - \gamma I
\]
\[
\dot C = \alpha SC
\]
\begin{enumerate}[label=(\alph*)]
\item Conservation? Reduce.
\item Find fixed points.
\item Compute Jacobian at disease-free FP.
\item Find eigenvalues. Define $R_0$ for each pathogen branch.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Conservation:}

\textbf{Step 1.} Sum all three equations:
\[
\dot S + \dot I + \dot C = (-\beta SI - \alpha SC + \gamma I) + (\beta SI - \gamma I) + (\alpha SC)
\]

\textbf{Step 2.} Group like terms:
\[
= (-\beta SI + \beta SI) + (-\alpha SC + \alpha SC) + (\gamma I - \gamma I) = 0 \checkmark
\]

\textbf{Step 3.} So $S + I + C = N$ is conserved. Use this to eliminate $S = N - I - C$ if desired. (For 3$\times$3 Jacobian we'll keep all three.)

\medskip\textbf{(b) Fixed points:}

\textbf{Step 1.} Set $\dot I = 0$:
\[
\beta SI - \gamma I = I(\beta S - \gamma) = 0 \thus I = 0 \text{ or } S = \gamma/\beta
\]

\textbf{Step 2.} Set $\dot C = 0$:
\[
\alpha SC = 0 \thus S = 0 \text{ or } C = 0
\]

\textbf{Step 3.} Disease-free: $I^* = 0$, $C^* = 0$, $S^* = N$. (Easy to verify $\dot S = 0$.)

\textbf{Step 4.} Endemic with $C^* = 0$ and $I^* > 0$: requires $S^* = \gamma/\beta$, then $I^* = N - \gamma/\beta$ (only positive if $R_0 = \beta N/\gamma > 1$).

\medskip\textbf{(c) Jacobian at disease-free FP $(S, I, C) = (N, 0, 0)$:}

\textbf{Step 1.} Define $f = \dot S$, $g = \dot I$, $h = \dot C$. Compute partials.

\textbf{Step 2.} For $f = -\beta SI - \alpha SC + \gamma I$:
\[
\partial f/\partial S = -\beta I - \alpha C
\]
\[
\partial f/\partial I = -\beta S + \gamma
\]
\[
\partial f/\partial C = -\alpha S
\]

\textbf{Step 3.} For $g = \beta SI - \gamma I$:
\[
\partial g/\partial S = \beta I, \quad \partial g/\partial I = \beta S - \gamma, \quad \partial g/\partial C = 0
\]

\textbf{Step 4.} For $h = \alpha SC$:
\[
\partial h/\partial S = \alpha C, \quad \partial h/\partial I = 0, \quad \partial h/\partial C = \alpha S
\]

\textbf{Step 5.} Build $J$ then evaluate at $(N, 0, 0)$:
\begin{itemize}
\item $\partial f/\partial S \to -\beta\cdot 0 - \alpha\cdot 0 = 0$
\item $\partial f/\partial I \to -\beta N + \gamma$
\item $\partial f/\partial C \to -\alpha N$
\item $\partial g/\partial S \to \beta\cdot 0 = 0$
\item $\partial g/\partial I \to \beta N - \gamma$
\item $\partial g/\partial C \to 0$
\item $\partial h/\partial S \to \alpha\cdot 0 = 0$
\item $\partial h/\partial I \to 0$
\item $\partial h/\partial C \to \alpha N$
\end{itemize}

\textbf{Step 6.} Final Jacobian:
\[
J^* = \begin{pmatrix} 0 & \gamma - \beta N & -\alpha N \\ 0 & \beta N - \gamma & 0 \\ 0 & 0 & \alpha N \end{pmatrix}
\]

\medskip\textbf{(d) Eigenvalues:}

\textbf{Step 1.} The first column is all zeros, so $\lambda_1 = 0$.

\textbf{Step 2.} The remaining 2$\times$2 sub-matrix is upper triangular:
\[
M = \begin{pmatrix} \beta N - \gamma & 0 \\ 0 & \alpha N \end{pmatrix}
\]
Eigenvalues are diagonal entries: $\lambda_2 = \beta N - \gamma$, $\lambda_3 = \alpha N$.

\textbf{Step 3.} Three eigenvalues:
\[
\boxed{\lambda_1 = 0, \quad \lambda_2 = \beta N - \gamma, \quad \lambda_3 = \alpha N}
\]

\textbf{Step 4.} Define $R_0$ for each transmission branch:
\[
R_0^{(I)} = \frac{\beta N}{\gamma} \quad\text{(infective branch)}
\]
For the chronic-carrier branch, since carriers never recover, the relevant ``$R_0$-like'' quantity is just $\alpha N$ (multiplied by an effective lifetime).

\textbf{Step 5.} Stability conditions:
\begin{itemize}
\item $\lambda_2 > 0 \iff R_0^{(I)} > 1 \iff$ infective branch invades.
\item $\lambda_3 > 0 \iff \alpha N > 0$ \textbf{always}.
\end{itemize}

\textbf{Step 6.} Conclusion: \textbf{the disease-free FP is always unstable} as long as $\alpha N > 0$ (any positive transmission rate). Once a chronic carrier exists, the system cannot return to disease-free.

\textbf{Why one $R_0$ is enough:} If $\alpha > 0$, $\lambda_3 > 0$ and instability is guaranteed regardless of $R_0^{(I)}$. So for stability we'd need BOTH $\alpha = 0$ (no chronic carrier transmission) AND $R_0^{(I)} < 1$. In practice, only $R_0^{(C)}$ matters for whether the system can be stable at all.
\end{solution}

\subsection{Problem 9.4: Multiple-Pack Hunting Variant (2024 Q1f)}

\begin{problem}
Now flip the pack-hunting scenario: 5 predators hunt and kill 100 prey at a time (like a dolphin pod consuming a school of fish).
\begin{enumerate}[label=(\alph*)]
\item What are the natural compartments?
\item Write the differential equations.
\item Interpret parameters vs standard LV.
\item Write $A$, find Tr, Det, eigenvalues, and comment on stability.
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Compartments:}

When 5 predators kill 100 prey at once, it's natural to track \textbf{schools of prey} $N_s$ (each school contains 100 prey) and \textbf{pods of predators} $P_p$ (each pod contains 5 predators). One predation event: 1 pod $\times$ 1 school $\to$ school eliminated, pod gets energy.

\medskip\textbf{(b) Equations:}

\[
\frac{dN_s}{dt} = r_s N_s - \bar\beta_s N_s P_p
\]
\[
\frac{dP_p}{dt} = \varepsilon\bar\beta_s N_s P_p - Z_p P_p
\]
Identical structure to standard LV, just in units of schools and pods.

\medskip\textbf{(c) Parameter interpretation:}

\begin{itemize}
\item $r_s$: school growth rate (new fish schools form). Same units as $r$ (time$^{-1}$).
\item $\bar\beta_s$: consumption rate per pod on schools. One pod kills one school at rate $\bar\beta_s$. In absolute fish numbers, $\bar\beta_s$ corresponds to eliminating 100 fish/encounter — much larger per-individual impact than standard LV.
\item $\varepsilon$: energy conversion efficiency (same as LV, dimensionless).
\item $Z_p$: pod mortality rate (same units as $Z$).
\end{itemize}

\textbf{Key differences in magnitudes:} Since each predation event now kills 100 prey (instead of 1), the effective consumption rate per individual predator is $\bar\beta_s/5$ (divided by pod size). However, in terms of the school-level model, $\bar\beta_s$ is comparable to standard $\bar\beta$.

\medskip\textbf{(d) Stability:}

\textbf{Step 1.} Same structure as standard LV:
\[
A = \begin{pmatrix} 0 & -\bar\beta_s \\ \varepsilon\bar\beta_s & 0 \end{pmatrix}
\]

\textbf{Step 2.} Tr $= 0$, Det $= \varepsilon\bar\beta_s^2 > 0$.

\textbf{Step 3.} Eigenvalues: $\lambda_\pm = \pm i\bar\beta_s\sqrt\varepsilon$. Pure imaginary $\Rightarrow$ neutral oscillations.

\textbf{Step 4.} \textbf{Same qualitative stability as standard LV}. However, because each predation event wipes out 100 prey, the oscillation amplitude in absolute prey numbers is much larger. The system reaches close to extinction of schools during the troughs. In practice, \textbf{more extinction risk} than standard LV.

\textbf{Compare 4-hunt-1 vs 5-hunt-100:}
\begin{itemize}
\item Both give Tr $= 0$, Det $> 0$, neutral oscillations.
\item Different magnitudes of $|\lambda|$ and oscillation amplitude.
\item Mathematical stability class: same. Practical biological stability: different.
\end{itemize}
\end{solution}

\subsection{Problem 9.5: Diagonalized Matrix and Eigenvalues (2019 Q2)}

\begin{problem}
For a 2$\times$2 matrix $A$ with eigenvalues $\lambda_1$, $\lambda_2$, the diagonalized version is:
\[
A' = \begin{pmatrix} \lambda_1 & 0 \\ 0 & \lambda_2 \end{pmatrix}
\]
\begin{enumerate}[label=(\alph*)]
\item Calculate $\text{Tr}(A')$ and $\text{Det}(A')$. Show these equal $\text{Tr}(A)$ and $\text{Det}(A)$.
\item Using these two equations, solve for $\lambda_1$ and $\lambda_2$ in terms of Tr and Det.
\item For a $3\times3$ matrix, how many independent equations do you need? What are they?
\end{enumerate}
\end{problem}

\begin{solution}
\textbf{(a) Tr and Det of $A'$:}

\textbf{Step 1.} Trace is the sum of diagonal entries:
\[
\text{Tr}(A') = \lambda_1 + \lambda_2
\]

\textbf{Step 2.} Determinant (diagonal matrix): product of diagonal entries:
\[
\text{Det}(A') = \lambda_1 \cdot \lambda_2 - 0\cdot 0 = \lambda_1\lambda_2
\]

\textbf{Step 3.} The trace and determinant are invariants under similarity transformations (diagonalization is a similarity transform $A' = P^{-1}AP$). Therefore:
\[
\text{Tr}(A) = \text{Tr}(A') = \lambda_1 + \lambda_2
\]
\[
\text{Det}(A) = \text{Det}(A') = \lambda_1\lambda_2
\]

\medskip\textbf{(b) Solve for $\lambda_1$, $\lambda_2$:}

\textbf{Step 1.} We have two equations:
\[
\lambda_1 + \lambda_2 = T \quad (T = \text{Tr})
\]
\[
\lambda_1\lambda_2 = D \quad (D = \text{Det})
\]

\textbf{Step 2.} These are Vieta's formulas for the roots of $\lambda^2 - T\lambda + D = 0$.

\textbf{Step 3.} Apply quadratic formula:
\[
\lambda = \frac{T \pm \sqrt{T^2 - 4D}}{2}
\]

This recovers the standard eigenvalue formula from two algebraic invariants.

\medskip\textbf{(c) 3$\times$3 matrix:}

\textbf{Step 1.} For 3$\times$3 with eigenvalues $\lambda_1, \lambda_2, \lambda_3$, the characteristic polynomial is degree 3:
\[
\lambda^3 - (\lambda_1+\lambda_2+\lambda_3)\lambda^2 + (\lambda_1\lambda_2+\lambda_1\lambda_3+\lambda_2\lambda_3)\lambda - \lambda_1\lambda_2\lambda_3 = 0
\]

\textbf{Step 2.} Three invariants needed:
\begin{itemize}
\item $I_1 = \lambda_1+\lambda_2+\lambda_3 = \text{Tr}(A)$
\item $I_2 = \lambda_1\lambda_2+\lambda_1\lambda_3+\lambda_2\lambda_3 = $ sum of 2$\times$2 principal minors
\item $I_3 = \lambda_1\lambda_2\lambda_3 = \text{Det}(A)$
\end{itemize}

\textbf{Step 3.} Trace and Det give $I_1$ and $I_3$; $I_2$ is a new quantity (not just Tr and Det).

\textbf{Step 4.} For $n\times n$: need $n$ invariants ($n$ coefficients of characteristic polynomial). Tr and Det always appear but for $n \geq 3$ you need $n-2$ additional ones.
\end{solution}

\newpage
%==================================================================
\section{Quick Reference: Common Tricks \& Patterns}
%==================================================================

\begin{tcolorbox}[colback=yellow!5,colframe=orange!70!black,title=\textbf{Stability Decision Tree}]
\begin{enumerate}
\item Find fixed point: set $d\vec X/dt = 0$.
\item If FP has any zero components $\to$ \textbf{must use Jacobian} (not interaction matrix).
\item If interactions are nonlinear (Type II, MM) $\to$ \textbf{must use Jacobian}.
\item Otherwise either method works at non-trivial FPs.
\item Compute eigenvalues from Tr and Det:
$\lambda_\pm = \frac{\text{Tr}}{2}\left[1 \pm \sqrt{1 - 4\text{Det}/\text{Tr}^2}\right]$.
\item Sign analysis:
\begin{itemize}
\item Tr $<0$, Det $>0$, Tr$^2 > 4\,$Det: stable node.
\item Tr $<0$, Det $>0$, Tr$^2 < 4\,$Det: stable spiral.
\item Det $<0$: saddle (unstable).
\item Tr $>0$, Det $>0$: unstable.
\item Tr $=0$, Det $>0$: neutral oscillations.
\end{itemize}
\end{enumerate}
\end{tcolorbox}

\begin{tcolorbox}[colback=blue!3,colframe=blue!70!black,title=\textbf{Common Modifications and Their Effects}]
\begin{tabular}{lll}
\textbf{Modification} & \textbf{Effect on Tr} & \textbf{Effect on stability}\\\hline
Logistic growth term & Adds $-r/K$ to diagonal & Stabilizing\\
Cannibalism & Adds $-\bar\beta_c$ to diagonal & Stabilizing\\
Type II response & N/A (changes structure) & Destabilizing\\
Predator carrying $K_P$ & Adds $-\varepsilon\bar\beta N^*/K_P$ & Stabilizing\\
Symbiosis $+\alpha N_1 N_2$ & Off-diagonal positive & Variable\\
Chronic carrier (no recovery) & Adds $+\alpha N^*$ & Destabilizing\\
\end{tabular}
\end{tcolorbox}

\begin{tcolorbox}[colback=red!3,colframe=red!70!black,title=\textbf{Tricks for Computing $R_0$}]
\begin{itemize}
\item For SIR: $R_0 = \beta N/\gamma$.
\item For SIR with deaths: $R_0 = \beta N/(\gamma + \mu)$.
\item For SI (no recovery): $R_0 \to \infty$ trivially (always spreads).
\item For multi-strain: each strain has its own $R_0$, computed as $(\text{transmission rate}\times \text{susceptible pool})/(\text{removal rate})$.
\item Vaccination with fraction $f$ effective: $R_0^\text{eff} = R_0(1 - f)$ (herd immunity threshold $f^* = 1 - 1/R_0$).
\end{itemize}
\end{tcolorbox}

\begin{tcolorbox}[colback=green!3,colframe=green!50!black,title=\textbf{Common Linear Spaces (Euler-Exact)}]
\begin{tabular}{ll}
\textbf{Equation} & \textbf{Linear-Euler Space}\\\hline
$\dot N = rN$ (exponential) & $\ln N$ vs $t$\\
$\dot N = rN/t$ (power) & $\ln N$ vs $\ln t$\\
$\dot N = r/t$ (inverse-time) & $N$ vs $\ln t$\\
$\dot N = rN^2$ (super-exp) & $1/N$ vs $t$\\
$\dot N = rN^p$ ($p\neq 1$) & $N^{1-p}$ vs $t$\\
$\dot N = rN(1-N/K)$ (logistic) & $\ln(N/(K-N))$ vs $t$\\
\end{tabular}
\end{tcolorbox}

\begin{tcolorbox}[colback=purple!3,colframe=purple!70!black,title=\textbf{When in Doubt on Exam}]
\begin{enumerate}
\item Always start by checking conservation. Reduces equations.
\item Always identify which fixed point matters (often disease-free for spread questions).
\item Always state whether to use $A$ matrix or $J^*$ \textbf{before} computing.
\item Compute Tr and Det \textbf{first}; they often answer the qualitative question.
\item For Taylor series questions: check if derivatives exist. For ``perfect space'' questions: ask what makes higher derivatives = 0.
\item For QSSA: identify the fast variable (intermediate complex) and set its derivative to 0.
\item For new modifications: ask ``does this add a diagonal term?'' (changes Tr) or ``does this break linearity?'' (kills $A$ approach).
\end{enumerate}
\end{tcolorbox}

\end{document}
